---
title: BayesGrove
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

[![CRAN status](https://img.shields.io/cran/v/release/bayesgrove)](https://cran.r-project.org/package/bayesgrove)
[![R-CMD-check](https://github.com/mscholz/bayesgrove/actions/workflows/R-CMD-check/badge.svg)](https://github.com/mscholz/bayesgrove/actions/workflows/R-CMD-check)
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT)

# BayesGrove: Graph-Based Bayesian Workflow Scaffolding

**BayesGrove** is an R-native, graph-based Bayesian workflow orchestrator. It helps you structure, track, and document your Bayesian analyses by separating the execution DAG (directed acyclic graph) from a rich decision-provenance layer.

Built on top of the pure functional graph engine [`dagriculture`](https://github.com/sims1253/dagriculture), BayesGrove provides explicit decision capture, branching, and reproducible workflow management.

## Key Features

- **Graph-Based Workflow**: Structure your analysis as a reproducible DAG of explicitly typed steps.
- **Decision Tracking (Provenance)**: Record all modeling choices, rationales, and literature references using structural "gates".
- **Deterministic Fingerprinting**: Precise, intent-based cache invalidation for partial reruns.
- **Workflow Guidance**: Derive obligations and suggested next actions from lightweight summaries via `bg_next_actions()`.
- **Reference-Semantic Handles**: Safely mutate and manage your workflow state using explicit S7-based project handles.

## Installation

```r
# Install dagriculture foundation first
# devtools::install_github("sims1253/dagriculture")

# Install BayesGrove from GitHub
# devtools::install_github("mscholz/bayesgrove")
```

## Quickstart

This minimal example shows how to initialize a project, build a simple graph, and execute it using the synchronous runner.

```{r quickstart}
library(bayesgrove)

# Initialize a project directory
project_path <- file.path(tempdir(), "bayesgrove-demo")
unlink(project_path, recursive = TRUE)

handle <- bg_init(
  project_path,
  project_name = "demo-project",
  workflow_packs = list("bayesguide.default_bayesian")
)

# Register simple mock executors for the demo
bg_register_node_kind(handle, "data", executor = function(node, inputs) {
  message("Loading data...")
  return(data.frame(x = 1:10))
})

bg_register_node_kind(handle, "model", executor = function(node, inputs) {
  message("Fitting model...")
  list(
    result = "mock_fit_result",
    summaries = list(list(
      summary_kind = "optimizer_diagnostics",
      passed = FALSE,
      severity = "warning",
      metrics = list(max_gradient = 0.1)
    ))
  )
})

bg_register_node_kind(handle, "compare", executor = function(node, inputs) {
  sprintf("Compared downstream result: %s", inputs[[1]])
})

# Build the workflow graph
n_data <- bg_add_node(handle, kind = "data", label = "Raw Data")
n_model <- bg_add_node(handle, kind = "model", label = "Baseline Model", inputs = n_data)
n_compare <- bg_add_node(
  handle,
  kind = "compare",
  label = "Diagnostic Comparison",
  inputs = n_model
)

# Create a decision gate
gate <- bg_add_gate(
  project = handle,
  from = n_data,
  to = n_model,
  prompt = "Does the data look ready for modeling?",
  options = c("yes", "no")
)

# Observe that the graph is blocked by the pending gate
plan <- bg_plan(handle)
print(plan$blocked)
print(bg_pending_gates(handle))

# Answer the gate explicitly
bg_answer_gate(
  project = handle,
  id = gate$id,
  choice = "yes",
  rationale = "No missing values found, ready to fit."
)

# Run the workflow
bg_run(handle, targets = n_model, mode = "sync")

# Ask the workflow protocol what needs attention next
next_steps <- bg_next_actions(handle)
next_steps$obligations[[1]]$kind

# Feed blocking obligations back into planning as external holds
held_plan <- bg_plan(
  handle,
  external_holds = next_steps$metadata$external_holds
)
held_plan$external_blocked
```

The final `bg_plan()` call is the key distinction: `held_plan$blocked` still
contains only structural blockers, while `held_plan$external_blocked` captures
workflow holds derived from summaries and obligations.

## Vignettes

For a detailed introduction, see our vignettes:
- `vignette("getting-started", package = "bayesgrove")`

## License

MIT © Maximilian Scholz
