---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bayesgrove

<!-- badges: start -->
[![R-CMD-check](https://github.com/sims1253/bayesgrove/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sims1253/bayesgrove/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/sims1253/bayesgrove/graph/badge.svg)](https://app.codecov.io/gh/sims1253/bayesgrove)
[![lint](https://github.com/sims1253/bayesgrove/actions/workflows/lint.yaml/badge.svg)](https://github.com/sims1253/bayesgrove/actions/workflows/lint.yaml)
[![format-check](https://github.com/sims1253/bayesgrove/actions/workflows/format-check.yaml/badge.svg)](https://github.com/sims1253/bayesgrove/actions/workflows/format-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Bayesian analysis is iterative by design. You fit a model, inspect diagnostics, revise your assumptions, branch into alternatives, compare candidates, and eventually commit to one path. The problem is that this process quietly disappears into unstructured scripts. Three months later—or when a collaborator picks up the project—the *why* behind every modeling decision is gone.

**bayesgrove** is a graph-based workflow orchestrator for iterative Bayesian analysis in R. While general-purpose pipelines like `targets` are built to skip up-to-date computations, bayesgrove is built to enforce statistical workflow. It separates the *execution graph* (the DAG of computations that run) from the *decision log* (why you ran them), and enforces a structured review loop: diagnostic warnings block downstream work, model criticism is recorded before repair, clean candidates are formally compared before one is accepted. `targets` doesn't know what a divergent transition is; bayesgrove does, and it will block your posterior predictive checks until you formally record how you plan to fix it.

## What you get

- **Decision provenance.** Attach explicit rationales to prior choices, reparametrizatons, data exclusions, or any other modeling decision—directly inside the workflow, not in a comment someone might miss.

- **Content-aware caching.** Each node carries a SHA-256 fingerprint of its inputs and parameters. Expensive MCMC fits are skipped when nothing upstream changed semantically. Change a prior and only the affected subtree reruns.

- **Enforced review.** The built-in workflow protocol derives obligations from the summaries your executors return. Warning diagnostics create a `review_computation_validity` obligation. Branch-scoped fit warnings also trigger `review_fit_criticism`. Once two candidates are clean, the protocol requires an explicit `compare_candidate_branches` step. Each branch then needs an `accept_or_reject_branch` disposition before the loop closes.

- **Branching with full lineage.** Fork from any node without losing the baseline. Branch goals, decisions, and diagnostic history travel with the branch. The lineage of any branch is always recoverable.

- **Reproducible handoffs.** Export the full decision tree and graph topology into a portable bundle or a structured markdown report with a single function call.

## Installation

```r
# install.packages("pak")
pak::pkg_install("sims1253/bayesgrove")
```

## A minimal example

The executor is a plain R function—you write the same `cmdstanr` code you
always would, and return a `summaries` list alongside the fit. *(Note: Future versions of bayesgrove will include built-in `cmdstanr` and `brms` executors to reduce boilerplate. This example builds one from scratch to show the mechanics).* bayesgrove reads
those summaries to derive workflow obligations, cache validity, and holds. The
rest of the machinery (DAG, fingerprinting, branching) is invisible until you
need it.

The example uses the eight-schools model in its centered parametrization, which
reliably produces divergent transitions and is the canonical motivation for
non-centered reparametrization.

```{r write-stan, include = FALSE}
stan_file <- tempfile(fileext = ".stan")
writeLines("
data {
  int<lower=1> J;
  vector[J] y;
  vector<lower=0>[J] sigma;
}
parameters {
  real mu;
  real<lower=0> tau;
  vector[J] theta;
}
model {
  mu    ~ normal(0, 5);
  tau   ~ normal(0, 1);
  theta ~ normal(mu, tau);
  y     ~ normal(theta, sigma);
}
", stan_file)
```

```stan
// eight_schools_centered.stan (saved as `stan_file` path)
data {
  int<lower=1> J;
  vector[J] y;
  vector<lower=0>[J] sigma;
}
parameters {
  real mu;
  real<lower=0> tau;
  vector[J] theta;     // centered — causes funnel geometry
}
model {
  mu    ~ normal(0, 5);
  tau   ~ normal(0, 1);
  theta ~ normal(mu, tau);
  y     ~ normal(theta, sigma);
}
```

```{r example, message = FALSE, warning = FALSE}
library(bayesgrove)

handle <- bg_init(
  path           = file.path(tempdir(), "bg-readme"),
  project_name   = "eight_schools",
  workflow_packs = list("bayesguide.default_bayesian")
)

bg_register_node_kind(handle, "data", executor = function(node, inputs) {
  list(
    J     = 8L,
    y     = c(28, 8, -3, 7, -1, 1, 18, 12),
    sigma = c(15, 10, 16, 11,  9, 11, 10, 18)
  )
})

bg_register_node_kind(handle, "fit", executor = function(node, inputs) {
  mod <- cmdstanr::cmdstan_model(node$params$stan_file, quiet = TRUE)
  fit <- mod$sample(
    data            = inputs[[1]],
    chains          = 2,
    parallel_chains = 2,
    iter_warmup     = 500,
    iter_sampling   = 500,
    seed            = 42L,
    refresh         = 0
  )

  diag     <- fit$diagnostic_summary(quiet = TRUE)
  rhat_max <- max(fit$summary()$rhat, na.rm = TRUE)
  n_div    <- sum(diag$num_divergent)
  passed   <- n_div == 0 && rhat_max < 1.01

  list(
    result    = fit,   # stored in the content-addressed cache
    summaries = list(list(
      summary_kind = "hmc_diagnostics",
      passed       = passed,
      severity     = if (passed) "ok" else "warning",
      metrics      = list(divergences = n_div, rhat_max = rhat_max)
    ))
  )
})

bg_register_node_kind(handle, "ppc", executor = function(node, inputs) {
  list(plot = "posterior_predictive_check")
})

n_data <- bg_add_node(handle, "data", label = "Eight schools data")
n_fit  <- bg_add_node(handle, "fit",  label = "Centered fit",
                       inputs = n_data,
                       params = list(stan_file = stan_file))
n_ppc  <- bg_add_node(handle, "ppc",  label = "Posterior check", inputs = n_fit)
```

```{r run, message = FALSE, warning = FALSE, results = "hide"}
bg_run(handle, targets = n_fit, mode = "sync")
```

The funnel geometry of the centered parametrization produces divergent
transitions. bayesgrove reads the `hmc_diagnostics` summary returned by the
executor and surfaces a blocking obligation automatically. We can inspect the summaries that triggered the hold:

```{r diagnostics}
summaries <- bg_read_summaries(handle)
summaries[[1]]$metrics
```

The posterior check is held from executing:

```{r protocol}
guide <- bg_next_actions(handle, scope = "project")

vapply(guide$obligations, `[[`, character(1), "kind")

held <- bg_plan(handle, external_holds = guide$metadata$external_holds)
length(held$external_blocked)
```

To resolve this, you record the criticism decision, branch from the problematic
fit, point the branch at a non-centered Stan file, and rerun. bayesgrove tracks
the lineage and keeps the original fit and its diagnostic history intact:

```{r write-ncp, include = FALSE}
ncp_stan_file <- tempfile(fileext = ".stan")
writeLines("
data {
  int<lower=1> J;
  vector[J] y;
  vector<lower=0>[J] sigma;
}
parameters {
  real mu;
  real<lower=0> tau;
  vector[J] theta_raw;
}
transformed parameters {
  vector[J] theta = mu + tau * theta_raw;
}
model {
  mu        ~ normal(0, 5);
  tau       ~ normal(0, 1);
  theta_raw ~ std_normal();
  y         ~ normal(theta, sigma);
}
", ncp_stan_file)
```

```r
# The protocol surfaced an obligation on the project because this is the main branch
bg_record_decision(
  handle,
  scope     = "project",
  prompt    = "How should the centered fit be handled?",
  choice    = "reparametrize_to_non_centered",
  rationale = "Divergences indicate the funnel geometry is poorly explored. Switch to non-centered.",
  kind      = "fit_criticism"
)

branch <- bg_branch_with_continuation(
  project            = handle,
  node_id            = n_fit,
  label              = "Non-centered reparametrization",
  continuation_kinds = "ppc"
)

# `ncp_stan_file` is the path to your non-centered Stan model
bg_update_node(handle, branch$branch$root_node_id,
               params = list(stan_file = ncp_stan_file))
bg_run(handle, targets = branch$branch$root_node_id, mode = "sync")
```

Once two clean candidates exist, the protocol requires an explicit comparison
and an accept/reject disposition for each branch before the loop closes:

```r
# We can now view the full lineage of both branches in the execution graph:
print(bg_read_graph(handle))
#> ── Execution Graph ──
#> 
#> └── Eight schools data <data> [new]
#>     ├── Centered fit <fit> [new]
#>     │   └── Posterior check <ppc> [new]
#>     └── Non-centered reparametrization <fit> [new]
#>         └── Posterior check (from branch) <ppc> [new]

# The protocol surfaces an explicit comparison step
bg_record_decision(
  handle,
  scope     = "project",
  prompt    = "Which branch should anchor the final report?",
  choice    = "prefer_non_centered_branch",
  rationale = "The non-centered branch resolves the divergences.",
  kind      = "model_comparison"
)

# Accept the non-centered branch
bg_record_decision(
  handle,
  scope     = branch$branch$branch_id,
  prompt    = "Should this branch be accepted?",
  choice    = "accept",
  rationale = "This is the clean branch preferred by the comparison step.",
  kind      = "branch_disposition"
)
```

The `guided-review-loop` vignette walks this full cycle end-to-end with
deterministic executors, covering branch lineage, stale evidence tracking,
comparison, and report export.

## Conceptual structure

bayesgrove organizes every analysis around three distinct layers:

| Layer | Contents |
|---|---|
| **Execution DAG** | Nodes (data, fit, check, compare), edges, cached results, content fingerprints |
| **Decision log** | Gates, explicit rationales, choices, timestamps, branch lineage |
| **Workflow protocol** | Obligations derived from executor summaries; external holds passed to the planner |

The execution DAG is built on
[`dagriculture`](https://github.com/sims1253/dagriculture), a purely functional
graph library. All workflow semantics—holds, branch scope, obligations, decision
provenance—are bayesgrove's own layer on top of the graph primitives.

## Vignettes

| Vignette | What it covers |
|---|---|
| `getting-started` | Full project lifecycle: gates, caching, branching, comparison, and disposition |
| `guided-review-loop` | Complete criticism-and-repair cycle with deterministic mock executors |
| `simulation-study` | Simulation-based workflow with confounded treatment and explicit gate decisions |
| `dagriculture-boundary` | How the graph primitives relate to workflow semantics |

```r
vignette("guided-review-loop", package = "bayesgrove")
```

## Background

The bayesgrove workflow is shaped by the iterative model-building perspective
described in Gelman et al. (2020, *Bayesian Workflow*,
[arXiv:2011.01808](https://arxiv.org/abs/2011.01808)) and the
simulation-based calibration approach of Talts et al. (2018,
[arXiv:1804.06788](https://arxiv.org/abs/1804.06788)). The obligation and
decision layers take direct inspiration from the practical need to record the
*why* behind each modeling step, not just the *what*—so that analysis lineage
is an artifact of the workflow, not an afterthought.
