---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BayesGrove

<!-- badges: start -->
[![R-CMD-check](https://github.com/sims1253/bayesgrove/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sims1253/bayesgrove/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**BayesGrove** is a graph-based Bayesian workflow engine for R. It helps you design, track, and document your statistical analyses by completely separating the *execution graph* (the code that runs) from the *decision provenance* (why you ran it).

Built on the purely functional `dagriculture` graph library, BayesGrove allows you to explicitly encode modeling decisions, enforce workflow checks, seamlessly cache intermediate computations like MCMC fits, and branch models asynchronously—all without losing the rationale behind your choices.

## Why BayesGrove?

Modern Bayesian analysis requires iterative model building, checking, and revision. Typically, this process gets lost in unstructured scripts or implicit notebook states. 

BayesGrove solves this by giving you:

1. **Explicit Provenance**: Record qualitative rationales (e.g., "Why did we use a heavy-tailed prior?") directly into the workflow using structural gates.
2. **Intent-Based Caching**: Skip redundant computations. If an upstream data step doesn't change semantically, your costly `cmdstanr` or `brms` fits won't rerun.
3. **Workflow Obligations**: The built-in protocol engine automatically flags diagnostic problems, asks for explicit fit criticism, requires formal comparison between clean candidates, and ends the loop with explicit branch acceptance or rejection.
4. **Reproducible Handoffs**: Export your entire decision tree and causal graph into a reproducible bundle or markdown report with one command.

For one deterministic end-to-end product loop, see the
`guided-review-loop` vignette. It walks through branch lineage, held
downstream work, explicit criticism and comparison decisions, branch
accept/reject dispositions, and report export without depending on a
heavyweight model fit.

## Installation

You can install the development version of BayesGrove (which includes its `dagriculture` foundation) via [`pak`](https://pak.r-lib.org/):

```r
# install.packages("pak")
pak::pkg_install("sims1253/bayesgrove")
```

## Example: Protocol-Driven Workflow

BayesGrove separates the execution graph from decision provenance. When computations produce diagnostic warnings, the protocol engine automatically creates blocking obligations and holds downstream work until you resolve them.

```{r example, message=FALSE, warning=FALSE}
library(bayesgrove)

# 1. Initialize a workflow workspace
project_path <- file.path(tempdir(), "bayesgrove-intro")
handle <- bg_init(
  path = project_path,
  project_name = "hierarchical_analysis",
  workflow_packs = list("bayesguide.default_bayesian")
)

# 2. Register executors that return diagnostic summaries
bg_register_node_kind(handle, "data_prep", executor = function(node, inputs) {
  data.frame(group = rep(1:5, each = 10), y = rnorm(50))
})

bg_register_node_kind(handle, "fit", executor = function(node, inputs) {
  # Mock fit that returns warning diagnostics (e.g., divergent transitions)
  list(
    status = "fit_completed",
    summaries = list(
      list(
        summary_kind = "hmc_diagnostics",
        severity = "warning",
        passed = FALSE,
        metrics = list(divergences = 15, rhat_max = 1.02)
      )
    )
  )
})

bg_register_node_kind(handle, "ppc", executor = function(node, inputs) {
  list(plot = "ppc_density_plot")
})

# 3. Construct the graph
n_data <- bg_add_node(handle, "data_prep", label = "Load Data")
n_fit <- bg_add_node(handle, "fit", label = "Hierarchical Fit", inputs = n_data)
n_ppc <- bg_add_node(handle, "ppc", label = "Posterior Check", inputs = n_fit)
```

Run the workflow. The fit completes but produces warning diagnostics:

```{r run}
bg_run(handle, mode = "sync")
```

The protocol engine detects the warning and creates a blocking obligation. Downstream work (the PPC) is held:

```{r status}
status <- bg_status(handle)
cat(sprintf("Workflow State: %s\n", status$workflow_state))
```

Inspect active obligations and suggested actions:

```{r guide}
actions <- bg_next_actions(handle)
cat(sprintf("Blocking obligations: %d\n", length(actions$obligations)))
if (length(actions$actions) > 0) {
  cat(sprintf("First suggested action: %s\n", actions$actions[[1]]$title))
}
```

To resolve the issue, you can branch and modify the fit with a non-centered parametrization. The interactive REPL provides a guided interface for this workflow:

```r
bg_repl(handle)
```

## Stronger Default Pack Loop

The built-in `bayesguide.default_bayesian` pack now enforces one narrow review loop:

1. warning or error summaries create a blocking `review_computation_validity` obligation;
2. branch-scoped fit warnings also create `review_fit_criticism`;
3. once two fits are clean, project scope creates `compare_candidate_branches`;
4. after a comparison exists, each candidate branch gets `accept_or_reject_branch` until you record `accept` or `reject`.

A concise end-to-end example looks like this:

```r
# After branching, modifying, and rerunning an alternative fit:
project_guide <- bg_next_actions(handle, scope = "project")
vapply(project_guide$obligations, `[[`, character(1), "kind")
#> "compare_candidate_branches"

# Create and run the comparison node, then record the project decision.
comparison_action <- Filter(
  function(x) identical(x$kind, "record_decision") &&
    identical(x$payload$decision_type, "model_comparison"),
  project_guide$actions
)[[1]]

bg_record_decision(
  handle,
  scope = "project",
  prompt = "Compare candidate branches",
  choice = "prefer_non_centered",
  rationale = "The alternative fit is cleaner and easier to trust.",
  kind = "model_comparison",
  metadata = comparison_action$payload[c(
    "fit_node_ids",
    "summary_ids",
    "candidate_signature",
    "comparison_signature",
    "comparison_context"
  )]
)

# Each candidate branch can then be explicitly accepted or rejected in
# branch scope with a `branch_disposition` decision.
```

## Built-in Templates

Common guided next steps are now exposed as first-class built-in templates:

- `diagnostic_check` creates a downstream `check` node from a fit
- `branch_comparison` creates a `compare` node across fit candidates
- `branch_and_modify_fit` wraps `bg_branch_with_continuation()`
- `review_decision` records explicit review decisions such as computation review, fit criticism, model comparison, and branch disposition

You can inspect the registry directly:

```r
bg_list_templates()
bg_list_templates("branch_comparison")
```

The interactive REPL uses the same registry when it executes template-backed actions.

## Interactive REPL

The REPL provides a guided loop for working through obligations:

- `status` - Show workflow state (blocked, idle, etc.)
- `guide` - See active obligations and suggested actions
- `actions` - List detailed actions with payloads
- `do <n>` - Execute action #n (branch and modify, record decision, etc.)
- `run` - Execute ready nodes
- `nodes` - View the execution graph with states
- `result <label>` - Inspect cached results and diagnostics
