---
title: "Eight-schools: a real cmdstanr workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Eight-schools: a real cmdstanr workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

<!-- The pasted output blocks below were captured from a real run via
     tools/precompute-vignettes.R (sampler progress lines and repeated
     warnings trimmed). Regenerate and refresh the blocks when bayesgrove or
     CmdStan changes; outputs are version-dependent.
     Last refreshed against bayesgrove 0.7.0 / CmdStan 2.39.0. -->

This vignette walks through one round of the review loop with real cmdstanr:
fit the eight-schools model in its centered parameterization, hit a
computation-review obligation, record the decision, branch to a non-centered
repair, and rerun. The chunks use `eval = FALSE` with output captured from a
real run, so the vignette builds without CmdStan on CI.

## Setup

```{r setup}
library(bayesgrove)

handle <- bg_init(
  path = file.path(tempdir(), "bg-eight-schools"),
  project_name = "Eight Schools"
)

# Activate the default review loop and the cmdstanr built-in executors.
bg_use_workflow_packs(handle, "bayesgrove.default_bayesian")
bg_use_cmdstanr(handle)
```

## Build the graph

```{r graph}
# Eight-schools data (Rubin, 1981; Gelman et al., 2013).
schools_data <- list(
  J = 8L,
  y = c(28, 8, -3, 7, -1, 1, 18, 12),
  sigma = c(15, 10, 16, 11, 9, 11, 10, 18)
)

# Node params are data, never code. To preserve R types that JSON serialization
# would mangle (e.g. integers), attach the data object via the content-addressed
# store with bg_set_node_data().
n_data <- bg_add_node(
  handle,
  kind = "stan_data",
  label = "Schools data"
)
bg_set_node_data(handle, n_data, schools_data)
```

The centered parameterization of the hierarchical model is a standard example
of funnel geometry: as `tau` approaches zero, the conditional distribution of
`theta` contracts sharply, and HMC produces divergent transitions
(Betancourt and Girolami, 2015). We write the Stan program to a file:

```{r stan-file}
stan_file <- tempfile(fileext = ".stan")
writeLines(c(
  "data { int<lower=1> J; vector[J] y; vector<lower=0>[J] sigma; }",
  "parameters { real mu; real<lower=0> tau; vector[J] theta; }",
  "model { mu ~ normal(0, 5); tau ~ cauchy(0, 5);",
  "        theta ~ normal(mu, tau); y ~ normal(theta, sigma); }"
), stan_file)

n_fit <- bg_add_node(
  handle,
  kind = "cmdstanr_fit",
  label = "Centered",
  inputs = n_data,
  params = list(
    stan_file = stan_file,
    chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    seed = 42
  )
)
```

## Run and hit the obligation

```{r run}
run <- bg_run(handle, targets = n_fit)
```

```
#> Starting run "run_179e87ae" with 1 node to execute.
#> Running node "node_9f8f0f45"...
#> Running node "node_55624d87"...
#> Running MCMC with 4 sequential chains...
#> All 4 chains finished successfully.
#> Warning: 81 of 4000 (2.0%) transitions ended with a divergence.
#> Warning: 2 of 4 chains had an E-BFMI less than 0.3.
```

The executor stores these diagnostics as a typed `hmc_diagnostics` summary:

```{r summaries}
summaries <- bg_read_summaries(handle)
summaries[[1]]$severity
#> [1] "error"
summaries[[1]]$metrics[c("divergences", "e_bfmi", "max_rhat")]
#> $divergences
#> [1] 81
#>
#> $e_bfmi
#> [1] 0.1574
#>
#> $max_rhat
#> [1] 1.0401
```

Because the summary carries `severity = "error"`, the default pack raises a
blocking `review_computation_validity` obligation and suggests two actions:
record the review decision, or branch and modify the fit.

```{r next-actions}
guide <- bg_next_actions(handle)
vapply(guide$obligations, `[[`, character(1), "kind")
#>                  obl_ef958c36
#> "review_computation_validity"
vapply(guide$actions, `[[`, character(1), "kind")
#>        act_2260dedb        act_663128f1
#>   "record_decision" "branch_and_modify"
```

## Record the review decision

The suggested `record_decision` action carries the decision kind and the ids
of the summaries under review in its payload. Recording a
`computation_review` decision with those ids resolves the obligation; a
decision of another kind, or one that does not reference the summaries, does
not.

```{r review-decision}
review <- Filter(
  function(a) identical(a$payload$decision_type, "computation_review"),
  guide$actions
)[[1]]

bg_record_decision(
  handle,
  scope = "project",
  prompt = "Is the centered fit acceptable for downstream use?",
  choice = "reject_and_reparametrize",
  rationale = paste(
    "81 of 4000 transitions diverged and E-BFMI is low: the sampler cannot",
    "explore the funnel between tau and theta. Switch to the non-centered",
    "parameterization."
  ),
  kind = "computation_review",
  metadata = review$payload[c("node_ids", "summary_ids")]
)

vapply(bg_next_actions(handle)$obligations, `[[`, character(1), "kind")
#> named character(0)
```

## Branch to the non-centered parameterization

The repair is a branch from the criticized fit. The branch keeps the original
fit and its diagnostics in the record; we give it an inferential goal (the
default pack asks for one on every active branch) and swap in the
non-centered program. `bg_update_node()` merges params, so the chains, iteration,
and seed settings carry over.

```{r branch}
branch <- bg_branch(handle, n_fit, label = "Non-centered")

bg_set_goal(
  project = handle,
  branch_id = branch$branch_id,
  kind = "latent_inference",
  label = "School effects, non-centered",
  rationale = "Same model; reparameterized so HMC can explore the posterior."
)

stan_file_nc <- tempfile(fileext = ".stan")
writeLines(c(
  "data { int<lower=1> J; vector[J] y; vector<lower=0>[J] sigma; }",
  "parameters { real mu; real<lower=0> tau; vector[J] theta_raw; }",
  "transformed parameters { vector[J] theta = mu + tau * theta_raw; }",
  "model { mu ~ normal(0, 5); tau ~ cauchy(0, 5);",
  "        theta_raw ~ normal(0, 1); y ~ normal(theta, sigma); }"
), stan_file_nc)

bg_update_node(
  handle,
  branch$root_node_id,
  params = list(stan_file = stan_file_nc)
)
```

## Rerun and close the loop

```{r rerun}
run2 <- bg_run(handle, targets = branch$root_node_id)
```

```
#> Starting run "run_a26583a8" with 1 node to execute.
#> Running node "node_4c430039"...
#> Running MCMC with 4 sequential chains...
#> All 4 chains finished successfully.
```

The non-centered fit samples cleanly. Both fits remain in the record: the
criticized centered fit with its error summary, and the repaired fit with a
clean one.

```{r rerun-summaries}
for (s in bg_read_summaries(handle, include_stale = FALSE)) {
  cat(s$node_id, "-", s$severity, "- divergences:", s$metrics$divergences, "\n")
}
#> node_55624d87 - error - divergences: 81
#> node_4c430039 - ok - divergences: 0
```

One obligation remains. Decisions are scope-specific (see
`vignette("concepts")`): the project-scope review recorded above does not
cover the new branch, so the branch asks for its own review of the centered
error summary in its lineage.

```{r branch-review}
branch_guide <- bg_next_actions(
  handle,
  scope = "branch",
  branch_id = branch$branch_id
)
vapply(branch_guide$obligations, `[[`, character(1), "kind")
#>                  obl_e02b096b
#> "review_computation_validity"

review_nc <- Filter(
  function(a) identical(a$payload$decision_type, "computation_review"),
  branch_guide$actions
)[[1]]

bg_record_decision(
  handle,
  scope = branch$branch_id,
  prompt = "Is the centered fit acceptable as this branch's baseline?",
  choice = "superseded_by_non_centered",
  rationale = paste(
    "The centered fit's divergences are the reason this branch exists;",
    "the non-centered fit replaces it."
  ),
  kind = "computation_review",
  metadata = review_nc$payload[c("node_ids", "summary_ids")]
)

bg_next_actions(handle)$obligations
#> named list()
```

No model comparison is needed here: the two fits target the same posterior,
and the reparameterization is a computational repair rather than a change of
model. For a loop with genuinely distinct candidates — where the default pack
requires an explicit comparison decision and an accept-or-reject disposition
for each branch — see `vignette("guided-review-loop")`.

## Export a report

```{r export}
print(bg_read_graph(handle))
#> ── Execution Graph ──
#>
#> └── Schools data <stan_data> [new]
#> ├── Centered <cmdstanr_fit> [new]
#> └── Non-centered <cmdstanr_fit> [new]

bg_export_report(handle, format = "md", path = "eight-schools-report.md")
```

The report captures the full trail: both fits, the diagnostics that fired,
the review decisions at each scope, and the branch lineage.
