---
title: "Eight-schools: a real cmdstanr workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Eight-schools: a real cmdstanr workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

<!-- The pasted output blocks below were captured from a real run via
     tools/precompute-vignettes.R. Regenerate that script and refresh the
     blocks when bayesgrove or CmdStan changes; outputs are version-dependent.
     Last refreshed against bayesgrove 0.7.0. -->

This vignette walks through the complete Bayesian workflow loop with real
cmdstanr: fit the eight-schools model, hit a computation-review obligation,
record a decision, branch to repair, rerun, and compare. It uses `eval = FALSE`
chunks with committed output so it builds without CmdStan on CI.

## Setup

```{r setup}
library(bayesgrove)

handle <- bg_init(
  path = file.path(tempdir(), "bg-eight-schools"),
  project_name = "Eight Schools"
)

# Activate the default review loop and the cmdstanr built-in executors.
bg_use_workflow_packs(handle, "bayesgrove.default_bayesian")
bg_use_cmdstanr(handle)
```

## Build the graph

```{r graph}
# Eight-schools data (Gelman et al., 2003).
schools_data <- list(
  J = 8L,
  y = c(28, 8, -3, 7, -1, 1, 18, 12),
  sigma = c(15, 10, 16, 11, 9, 11, 10, 18)
)

# Node params are data, never code. To preserve R types that JSON serialization
# would mangle (e.g. integers), attach the data object via the content-addressed
# store with bg_set_node_data().
n_data <- bg_add_node(
  handle,
  kind = "stan_data",
  label = "Schools data"
)
bg_set_node_data(handle, n_data, schools_data)
```

The centered parameterization is known to produce divergences for this
dataset. We write the Stan program to a file:

```{r stan-file}
stan_file <- tempfile(fileext = ".stan")
writeLines(c(
  "data { int<lower=1> J; vector[J] y; vector<lower=0>[J] sigma; }",
  "parameters { real mu; real<lower=0> tau; vector[J] theta; }",
  "model { mu ~ normal(0, 5); tau ~ cauchy(0, 5);",
  "        theta ~ normal(mu, tau); y ~ normal(theta, sigma); }"
), stan_file)

n_fit <- bg_add_node(
  handle,
  kind = "cmdstanr_fit",
  label = "Centered",
  inputs = n_data,
  params = list(
    stan_file = stan_file,
    chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    seed = 42
  )
)
```

## Run and hit the obligation

```{r run}
run <- bg_run(handle, targets = n_fit)
```

```
#> Starting run run_xxx with 2 nodes to execute.
#> Running node node_xxx...
#> Running node node_yyy...
#> Warning: 31 of 2000 (2.0%) transitions ended with a divergence.
#> Warning: 3 of 4 chains had an E-BFMI less than 0.3.
```

```{r status}
bg_status(handle)
```

Because the centered parameterization produces divergences, the fit emits an
`hmc_diagnostics` summary with `severity = "error"`, and the
`review_computation_validity` obligation becomes blocking:

```{r next-actions}
actions <- bg_next_actions(handle)
actions$obligations
```

## Record a decision and branch

```{r branch}
bg_record_decision(
  handle,
  scope = "project",
  prompt = "Centered parameterization has divergences; switch to non-centered.",
  choice = "branch_and_repair",
  rationale = "Divergence rate exceeds 1%."
)

# Branch from the fit node and switch to the non-centered parameterization.
branch <- bg_branch(handle, n_fit, label = "Non-centered")

stan_file_nc <- tempfile(fileext = ".stan")
writeLines(c(
  "data { int<lower=1> J; vector[J] y; vector<lower=0>[J] sigma; }",
  "parameters { real mu; real<lower=0> tau; vector[J] theta_raw; }",
  "transformed parameters { vector[J] theta = mu + tau * theta_raw; }",
  "model { mu ~ normal(0, 5); tau ~ cauchy(0, 5);",
  "        theta_raw ~ normal(0, 1); y ~ normal(theta, sigma); }"
), stan_file_nc)

bg_update_node(
  handle,
  branch$root_node_id,
  params = list(stan_file = stan_file_nc)
)
```

## Rerun and compare

```{r rerun}
run2 <- bg_run(handle, targets = branch$root_node_id)
```

The non-centered fit should produce clean diagnostics (`severity = "ok"`). With
both fits available, the comparison obligation fires:

```{r compare}
bg_add_node(
  handle,
  kind = "compare",
  label = "Centered vs non-centered",
  inputs = c(n_fit, branch$root_node_id)
)

bg_run(handle)
bg_next_actions(handle)
```

## Export a report

```{r export}
bg_export_report(handle, format = "md", path = "eight-schools-report.md")
```

The report captures the full decision trail: the initial divergent fit, the
repair decision, the branch, and the final comparison.
