---
title: "Extending bayesgrove"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending bayesgrove}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Why this vignette exists

bayesgrove keeps the runtime boundary small on purpose. You bring your own node
kinds, executors, and domain conventions. bayesgrove handles the graph, cache,
decision log, and workflow protocol.

This vignette uses the following vocabulary:

| Term | Meaning in bayesgrove today |
|---|---|
| Node set | A group of node kinds plus executors you register on a handle |
| Ruleset | A chosen set of `workflow_packs` plus the summary vocabulary they consume |
| Domain module | An R helper or package that bundles node registration, pack selection, and project conventions |
| Plugin | A built-in executor set (e.g. `bg_use_cmdstanr()`) |
| Extension | Any combination of the four items above |

The stable public hooks are node-kind registration, built-in executor
activation (`bg_use_cmdstanr()` / `bg_use_brms()`), and summary emission from
executors. Workflow-pack selection is also public. Authoring brand new
workflow-pack providers is still an internal extension surface, so external
packages should currently target the built-in summary contracts instead of
reaching into unexported provider internals.

# Activate the built-in workflow packs you need

Projects now start empty by default. If your workflow needs the built-in
review vocabulary, activate and persist the packs you want after creating the
project:

```{r setup-packs}
library(bayesgrove)

project_root <- file.path(tempdir(), "bg-extensions")
unlink(project_root, recursive = TRUE, force = TRUE)

handle <- bg_init(
  path = project_root,
  project_name = "Extensions"
)

bg_use_workflow_packs(
  handle,
  c(
    "bayesgrove.default_bayesian",
    "bayesgrove.process_guidance",
    "bayesgrove.model_taxonomy",
    "bayesgrove.prior_workflow",
    "bayesgrove.model_checks",
    "bayesgrove.model_selection",
    "bayesgrove.stan_workflow",
    "bayesgrove.causal_dagitty"
  )
)

vapply(bg_workflow_packs(handle), function(pack) pack$pack_id, character(1))
```

Those packs add review vocabulary without changing the graph engine:

- `bayesgrove.default_bayesian`: computation review, fit criticism,
  candidate comparison, and branch disposition.
- `bayesgrove.process_guidance`: workflow preflight, iterative repair notes,
  and out-of-sample stability review.
- `bayesgrove.model_taxonomy`: PAD classification and utility-tradeoff review.
- `bayesgrove.prior_workflow`: prior rationale and prior predictive review.
- `bayesgrove.model_checks`: posterior predictive checks and, for
  `latent_inference` branches, SBC and LOO-PIT calibration review.
- `bayesgrove.model_selection`: model-comparison evidence from fresh
  comparison summaries, with optional stacking weights.
- `bayesgrove.stan_workflow`: a bundled Stan-oriented ruleset covering the
  default loop, process guidance, taxonomy, predictive checks, model
  selection, Stan diagnostics, and projection review.
- `bayesgrove.causal_dagitty`: DAG-backed adjustment review, implication
  review, and causal selection contracts for `latent_inference` branches.

The scope matters. `bayesgrove.model_selection` only acts at project scope
when there are at least two clean fit candidates. The taxonomy and causal packs
are branch-scoped, and they do not fire until the branch has the right goal
context. In particular, `bayesgrove.causal_dagitty` becomes stronger after a
reviewed `causal_selection_contract`: formulas and projection-based selection
can then be constrained to keep required identification terms and exclude
forbidden controls.

# Build a node set

In practice, a node set is usually a helper that registers several node kinds
for one modeling domain. The only hard contract is that each executor receives
`(node, inputs)` and returns plain R data.

```{r node-set, eval = FALSE}
register_normal_model_nodes <- function(handle) {
  bg_register_node_kind(handle, "study_data", executor = function(node, inputs) {
    data.frame(y = rnorm(50), x = rnorm(50))
  })

  bg_register_node_kind(handle, "prior_note", executor = function(node, inputs) {
    list(
      result = node$params,
      summaries = list(list(
        summary_kind = "prior_spec",
        passed = TRUE,
        severity = "ok",
        metrics = list(scale = node$params$scale)
      ))
    )
  })

  bg_register_node_kind(handle, "fit_normal", executor = function(node, inputs) {
    data <- inputs[[1]]

    list(
      result = list(mean_y = mean(data$y)),
      summaries = list(
        list(
          summary_kind = "prior_predictive_check",
          passed = TRUE,
          severity = "ok",
          metrics = list(simulated_mean = 0.0)
        ),
        list(
          summary_kind = "posterior_predictive_check",
          passed = TRUE,
          severity = "ok",
          metrics = list(ppc_p_value = 0.46)
        ),
        list(
          summary_kind = "pad_annotation",
          passed = TRUE,
          severity = "ok",
          metrics = list(
            model_class = "PAD",
            primary_utility = "predictive_performance"
          )
        )
      )
    )
  })
}
```

Two practical rules make these node sets easy to compose:

1. Keep executors domain-specific and return plain data.
2. Emit summaries as plain-data records so workflow packs can react to them.

## Opening a project never runs project-supplied code

Executors are R functions, and a project's runtime manifest stores their
source text so the project can be reopened later. For safety, `bg_open()`
restores persisted node kinds **structurally only** — the kind name, input
contracts, and stored executor source are kept as inert metadata, but the
source is never evaluated. Opening a project directory from an untrusted
source therefore executes nothing.

To restore executable executors after opening, call
`bg_restore_executors()` explicitly:

```{r restore, eval = FALSE}
handle <- bg_open("path/to/project")

# Default: print the stored source for each kind and abort, so you can
# review what would run before consenting.
bg_restore_executors(handle)

# Once you have reviewed the source, opt in:
bg_restore_executors(handle, trust = TRUE)
```

With `trust = FALSE` (the default) the stored source for each kind is
printed and the call aborts. With `trust = TRUE` each source is evaluated
in a child of the global environment (not inside the bayesgrove
namespace), and a warning reminds you that closures capturing the original
environment are not restored. Alternatively, re-register executors with
`bg_register_node_kind()`.

Built-in executors (shipped with the package) are persisted as stable
references and restore safely without this opt-in step.

### Params are data, never code

The same trust boundary applies to node `params`. A node's params carry
**data** only — never R source. There is no params channel that evaluates
arbitrary code at run time, so a hostile `graph.json` cannot smuggle code in
through a `data_fn_source` or `check_fn_source` param (those have been
removed). To attach a data object that JSON serialization would mangle (for
example Stan's typed integers), use `bg_set_node_data()`, which stores the
object in the content-addressed store and records a `data_ref` param pointing
at it:

```{r set-data, eval = FALSE}
n_data <- bg_add_node(handle, kind = "stan_data", label = "Data")
bg_set_node_data(handle, n_data, list(N = 10L, y = as.integer(y)))
```

Code enters a project only through the executor gate above — by registering a
node kind in-session with `bg_register_node_kind()`, or by consenting with
`bg_restore_executors(trust = TRUE)`.

# Emit summaries that workflow packs understand

The optional packs are summary-driven. They do not require special node kinds;
they only care about summary records attached to fresh results.

| Summary kind | Consumed by | Typical use |
|---|---|---|
| `prior_spec` | `bayesgrove.prior_workflow` | Record the current prior specification |
| `prior_predictive_check` | `bayesgrove.prior_workflow` | Review whether priors generate plausible observables |
| `posterior_predictive_check` | `bayesgrove.model_checks` | Review fit-to-data mismatch after conditioning |
| `loo_pit_calibration` | `bayesgrove.model_checks` | Review leave-one-out predictive calibration |
| `sbc_result` | `bayesgrove.model_checks` | Review parameter recoverability and calibration for `latent_inference` branches |
| `comparison_results` / `model_comparison` / `stacking_weights` | `bayesgrove.model_selection` | Compare clean candidates and review weighting evidence |
| `projpred_selection` / `projection_predictive_selection` | `bayesgrove.stan_workflow` | Review submodel compression and refit plans |
| `dagitty_adjustment` | `bayesgrove.causal_dagitty` | Review DAG-derived adjustment sets |
| `dagitty_implications` | `bayesgrove.causal_dagitty` | Review implied independencies and falsification targets |
| `causal_selection_contract` | `bayesgrove.causal_dagitty` | Lock required terms, exclude forbidden controls, and rank admissible precision candidates |
| `pad_annotation` | `bayesgrove.model_taxonomy` | Label a branch with PAD class and utility dimensions |

# Causal selection contracts and constrained projpred workflows

The new causal-DAG path is meant to guide model construction, not only review
it after the fact. A causal branch can now emit a `causal_selection_contract`
summary with fields such as:

- `required_terms` for identification-critical covariates,
- `forbidden_terms` for colliders, post-treatment variables, or other blocked
  controls,
- `ranked_candidate_terms` for the admissible precision path from simpler to
  richer models, and
- optionally `allowed_formulas` when the DAG layer wants to enumerate exact
  permitted RHS patterns.

Once that summary is reviewed, `bayesgrove.causal_dagitty` can enforce formula
consistency and keep downstream work on hold until the branch's formulas obey
the contract. `bayesgrove.stan_workflow` also forwards the same causal payload
into projection-predictive review so a `projpred`-style node can search only
inside the admissible ranked candidate set.

```{r causal-contract, eval = FALSE}
bg_register_node_kind(handle, "causal_selection_contract", executor = function(node, inputs) {
  list(
    summaries = list(list(
      summary_kind = "causal_selection_contract",
      passed = TRUE,
      severity = "ok",
      metrics = list(
        required_terms = c("treatment", "z"),
        forbidden_terms = "post_treatment",
        ranked_candidate_terms = c("w", "x")
      )
    ))
  )
})
```

That contract keeps the Bayesian-workflow iteration spirit intact: start with
the smallest causally valid formula, then add ranked admissible terms as the
workflow pursues more precision or better predictive compression.

Each summary should stay plain and serializable. A good pattern is:

- `summary_kind` for dispatch,
- `passed` and `severity` for review urgency,
- `metrics` for compact numeric or categorical evidence,
- extra fields only when they remain stable plain data.

# Package a domain module

A domain module is just a reusable setup layer around `bg_init()` and your node
registration helpers. This is the easiest way to publish a project-specific
extension package.

```{r domain-module, eval = FALSE}
init_normal_workflow <- function(path) {
  handle <- bg_init(
    path = path,
    project_name = "normal_workflow"
  )

  bg_use_workflow_packs(
    handle,
    c(
      "bayesgrove.default_bayesian",
      "bayesgrove.process_guidance",
      "bayesgrove.model_taxonomy",
      "bayesgrove.prior_workflow",
      "bayesgrove.model_checks",
      "bayesgrove.model_selection",
      "bayesgrove.stan_workflow"
    )
  )

  register_normal_model_nodes(handle)
  handle
}
```

That helper can live in your analysis package, a lab-internal utility package,
or even a plain project script. The important part is that the node set and the
selected workflow packs evolve together.

# Built-in cmdstanr and brms executors

For real model fitting, bayesgrove ships built-in executors that wrap
cmdstanr and brms and compute diagnostics themselves (divergences, R-hat,
ESS, Pareto-k, posterior-predictive p-values). Activate them with:

```{r builtin-executors, eval = FALSE}
bg_use_cmdstanr(handle)
bg_use_brms(handle)
```

These register node kinds (`stan_data`, `cmdstanr_fit`, `brms_fit`, `loo`,
`compare`, `ppc`, ...) whose executors emit the validated summary kinds the
workflow packs consume. You do not need to hand-roll diagnostics: a fit
through `bg_use_cmdstanr()` emits an `hmc_diagnostics` summary automatically,
and a blocking `review_computation_validity` obligation follows when
diagnostics are poor.

Custom executors are still possible via `bg_register_node_kind()`; but the
built-in executors are the documented happy path for cmdstanr and brms.


# What a ruleset means right now

Today, a practical "ruleset" in bayesgrove means two things together:

1. the `workflow_packs` you activate in `bg_init()`, and
2. the summary kinds your executors emit for those packs to interpret.

That path is public and stable enough to build around. By contrast,
registering entirely new workflow-pack providers from an external package is
not yet a public API. If you need custom review logic today, prefer one of
these approaches:

- emit the existing summary vocabulary and reuse the built-in packs,
- wrap a project-specific decision policy around `bg_next_actions()`, or
- pin to internal helpers only if you are comfortable tracking internal API
  changes.

# Recommended extension shape

For most projects, the easiest sustainable pattern is:

1. create a domain module that initializes the right workflow packs,
2. register a small node set for your modeling tasks,
3. use built-in backend plugins where possible,
4. emit plain-data summaries that match the review vocabulary you want, and
5. keep project-specific heuristics outside the core runtime.

That keeps bayesgrove responsible for provenance, caching, branching, and
review state while your package owns the domain logic.
