---
title: "Extending bayesgrove"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending bayesgrove}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Register your own node kinds and executors. bayesgrove handles the graph, cache,
decision log, and workflow protocol.

| Term | Meaning in bayesgrove |
|---|---|
| Node set | A group of node kinds plus executors you register on a handle |
| Ruleset | A chosen set of `workflow_packs` plus the summary vocabulary they consume |
| Domain module | An R helper or package that bundles node registration, pack selection, and project conventions |
| Plugin | A built-in executor set (e.g. `bg_use_cmdstanr()`) |
| Extension | Any combination of the four items above |

The stable public hooks are node-kind registration, built-in executor
activation (`bg_use_cmdstanr()` / `bg_use_brms()`), and summary emission from
executors. Workflow-pack selection is also public. Custom workflow-pack
providers use internal APIs. External packages should emit the built-in
summary kinds rather than call unexported provider functions.

# Select workflow packs

Projects start empty by default. If your workflow needs the built-in
review vocabulary, activate and persist the packs you want after creating the
project:

```{r setup-packs}
library(bayesgrove)

project_root <- file.path(tempdir(), "bg-extensions")
unlink(project_root, recursive = TRUE, force = TRUE)

handle <- bg_init(
  path = project_root,
  project_name = "Extensions"
)

bg_use_workflow_packs(
  handle,
  c(
    "bayesgrove.default_bayesian",
    "bayesgrove.process_guidance",
    "bayesgrove.model_taxonomy",
    "bayesgrove.prior_workflow",
    "bayesgrove.model_checks",
    "bayesgrove.model_selection",
    "bayesgrove.stan_workflow",
    "bayesgrove.causal_dagitty"
  )
)

vapply(bg_workflow_packs(handle), function(pack) pack$pack_id, character(1))
```

Those packs add review vocabulary without changing the graph engine:

- `bayesgrove.default_bayesian`: computation review, fit criticism,
  candidate comparison, and branch disposition.
- `bayesgrove.process_guidance`: workflow preflight, iterative repair notes,
  and out-of-sample stability review.
- `bayesgrove.model_taxonomy`: PAD classification and utility-tradeoff review.
- `bayesgrove.prior_workflow`: prior rationale and prior predictive review.
- `bayesgrove.model_checks`: posterior predictive checks and, for
  `latent_inference` branches, SBC and LOO-PIT calibration review.
- `bayesgrove.model_selection`: model-comparison evidence from fresh
  comparison summaries, with optional stacking weights.
- `bayesgrove.stan_workflow`: a bundled Stan-oriented ruleset covering the
  default loop, process guidance, taxonomy, predictive checks, model
  selection, Stan diagnostics, and projection review.
- `bayesgrove.causal_dagitty`: DAG-backed adjustment review, implication
  review, and causal selection contracts for `latent_inference` branches.

`bayesgrove.model_selection` only acts at project scope
when there are at least two clean fit candidates. The taxonomy and causal packs
apply to branches with the relevant goal. After review of a
`causal_selection_contract`, `bayesgrove.causal_dagitty` can constrain formulas
and projection-based selection to retain required identification terms and
exclude forbidden controls.

# Build a node set

A node set is usually a helper that registers several node kinds for one
modeling domain. Each executor receives
`(node, inputs)` and returns plain R data.

```{r node-set, eval = FALSE}
register_normal_model_nodes <- function(handle) {
  bg_register_node_kind(handle, "study_data", executor = function(node, inputs) {
    data.frame(y = rnorm(50), x = rnorm(50))
  })

  bg_register_node_kind(handle, "prior_note", executor = function(node, inputs) {
    list(
      result = node$params,
      summaries = list(list(
        summary_kind = "prior_spec",
        passed = TRUE,
        severity = "ok",
        metrics = list(scale = node$params$scale)
      ))
    )
  })

  bg_register_node_kind(handle, "fit_normal", executor = function(node, inputs) {
    data <- inputs[[1]]

    list(
      result = list(mean_y = mean(data$y)),
      summaries = list(
        list(
          summary_kind = "prior_predictive_check",
          passed = TRUE,
          severity = "ok",
          metrics = list(simulated_mean = 0.0)
        ),
        list(
          summary_kind = "posterior_predictive_check",
          passed = TRUE,
          severity = "ok",
          metrics = list(ppc_p_value = 0.46)
        ),
        list(
          summary_kind = "pad_annotation",
          passed = TRUE,
          severity = "ok",
          metrics = list(
            model_class = "PAD",
            primary_utility = "predictive_performance"
          )
        )
      )
    )
  })
}
```

For node sets that can be used together:

1. Keep executors domain-specific and return plain data.
2. Emit summaries as plain-data records so workflow packs can react to them.

## Opening a project never runs project-supplied code

Executors are R functions, and a project's runtime manifest stores their
source text so the project can be reopened later. For safety, `bg_open()`
restores the kind name, input contracts, and executor source as metadata.
It does not evaluate the saved source, including when opening a project
directory from an untrusted source.

To restore executable executors after opening, call
`bg_restore_executors()` explicitly:

```{r restore, eval = FALSE}
handle <- bg_open("path/to/project")

# Default: print the stored source for each kind and abort, so you can
# review what would run before consenting.
bg_restore_executors(handle)

# Once you have reviewed the source, opt in:
bg_restore_executors(handle, trust = TRUE)
```

With `trust = FALSE` (the default) the stored source for each kind is
printed and the call aborts. With `trust = TRUE` each source is evaluated
in a child of the global environment (not inside the bayesgrove
namespace), and a warning reminds you that closures capturing the original
environment are not restored. Alternatively, re-register executors with
`bg_register_node_kind()`.

Built-in executors (shipped with the package) are persisted as stable
references and restore safely without this opt-in step.

### Attach data to nodes

Node `params` contain data, not executable R source. bayesgrove does not
evaluate R code from params, including `data_fn_source` or `check_fn_source`.
To preserve R types that JSON serialization can change, such as Stan's typed
integers, attach the object with `bg_set_node_data()`. It saves the object in
the content-addressed store and sets a `data_ref` param:

```{r set-data, eval = FALSE}
n_data <- bg_add_node(handle, kind = "stan_data", label = "Data")
bg_set_node_data(handle, n_data, list(N = 10L, y = as.integer(y)))
```

Register executable R functions with `bg_register_node_kind()`, or restore
saved executors with `bg_restore_executors(trust = TRUE)` after reviewing them.

# Emit diagnostic summaries

The optional packs are summary-driven. They do not require special node kinds;
they use summary records attached to fresh results.

| Summary kind | Consumed by | Typical use |
|---|---|---|
| `prior_spec` | `bayesgrove.prior_workflow` | Record the current prior specification |
| `prior_predictive_check` | `bayesgrove.prior_workflow` | Review whether priors generate plausible observables |
| `posterior_predictive_check` | `bayesgrove.model_checks` | Review fit-to-data mismatch after conditioning |
| `loo_pit_calibration` | `bayesgrove.model_checks` | Review leave-one-out predictive calibration |
| `sbc_result` | `bayesgrove.model_checks` | Review parameter recoverability and calibration for `latent_inference` branches |
| `comparison_results` / `model_comparison` / `stacking_weights` | `bayesgrove.model_selection` | Compare clean candidates and review weighting evidence |
| `projpred_selection` / `projection_predictive_selection` | `bayesgrove.stan_workflow` | Review submodel compression and refit plans |
| `dagitty_adjustment` | `bayesgrove.causal_dagitty` | Review DAG-derived adjustment sets |
| `dagitty_implications` | `bayesgrove.causal_dagitty` | Review implied independencies and falsification targets |
| `causal_selection_contract` | `bayesgrove.causal_dagitty` | Lock required terms, exclude forbidden controls, and rank admissible precision candidates |
| `pad_annotation` | `bayesgrove.model_taxonomy` | Label a branch with PAD class and utility dimensions |

# Causal selection contracts and constrained projpred workflows

A causal branch can emit a `causal_selection_contract` summary to constrain
model construction. Its fields include:

- `required_terms` for identification-critical covariates,
- `forbidden_terms` for colliders, post-treatment variables, or other blocked
  controls,
- `ranked_candidate_terms` for the admissible precision path from simpler to
  larger models, and
- optionally `allowed_formulas` to enumerate exact
  permitted RHS patterns.

Once that summary is reviewed, `bayesgrove.causal_dagitty` can enforce formula
consistency and keep downstream work on hold until the branch's formulas obey
the contract. `bayesgrove.stan_workflow` also forwards the same causal payload
into projection-predictive review so a `projpred`-style node can search only
inside the admissible ranked candidate set.

```{r causal-contract, eval = FALSE}
bg_register_node_kind(handle, "causal_selection_contract", executor = function(node, inputs) {
  list(
    summaries = list(list(
      summary_kind = "causal_selection_contract",
      passed = TRUE,
      severity = "ok",
      metrics = list(
        required_terms = c("treatment", "z"),
        forbidden_terms = "post_treatment",
        ranked_candidate_terms = c("w", "x")
      )
    ))
  )
})
```

Start with the smallest causally valid formula. Add permitted terms in the
specified order when evaluating precision. Restrict projection-based
selection to the permitted candidates.

Use serializable data in each summary:

- `summary_kind` for dispatch,
- `passed` and `severity` for review urgency,
- `metrics` for compact numeric or categorical evidence,
- extra fields only when they remain stable plain data.

# Package a domain module

A domain module combines `bg_init()` with node-registration helpers in a
reusable setup function. You can distribute it as an extension package.

```{r domain-module, eval = FALSE}
init_normal_workflow <- function(path) {
  handle <- bg_init(
    path = path,
    project_name = "normal_workflow"
  )

  bg_use_workflow_packs(
    handle,
    c(
      "bayesgrove.default_bayesian",
      "bayesgrove.process_guidance",
      "bayesgrove.model_taxonomy",
      "bayesgrove.prior_workflow",
      "bayesgrove.model_checks",
      "bayesgrove.model_selection",
      "bayesgrove.stan_workflow"
    )
  )

  register_normal_model_nodes(handle)
  handle
}
```

Put the helper in an analysis package, a lab utility package, or a project
script. Update the node set and its selected workflow packs together.

# Built-in cmdstanr and brms executors

bayesgrove includes executors that call
cmdstanr and brms and compute diagnostics themselves (divergences, R-hat,
ESS, Pareto-k, posterior-predictive p-values). Activate them with:

```{r builtin-executors, eval = FALSE}
bg_use_cmdstanr(handle)
bg_use_brms(handle)
```

These register node kinds (`stan_data`, `cmdstanr_fit`, `brms_fit`, `loo`,
`compare`, `ppc`, ...) whose executors emit the validated summary kinds the
workflow packs consume. A `cmdstanr_fit` node emits an `hmc_diagnostics`
summary. With the default review pack active, warning or error diagnostics
raise a blocking `review_computation_validity` obligation.

Use the built-in executors for cmdstanr and brms, or register a custom
executor with `bg_register_node_kind()`.


# Rulesets

A ruleset combines:

1. the `workflow_packs` you activate in `bg_init()`, and
2. the summary kinds your executors emit for those packs to interpret.

Registering new workflow-pack providers from an external package is
not a public API. For custom review logic, use one of
these approaches:

- emit the existing summary vocabulary and reuse the built-in packs,
- wrap a project-specific decision policy around `bg_next_actions()`, or
- pin to internal helpers only if you are comfortable tracking internal API
  changes.
