---
title: "Getting Started with BayesGrove"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with BayesGrove}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

**BayesGrove** is a graph-based Bayesian workflow orchestrator designed to track the iterative nature of model building. It separates your analysis into two distinct layers:
1. **The Execution Graph (DAG)**: Built using `dagriculture`, representing the step-by-step transformations and computations (data cleaning, compiling, fitting, diagnosing).
2. **The Decision Layer**: An annotation overlay that captures the provenance of your choices (why did you change the prior? why did you filter outliers?) via decision gates.

This separation ensures that your computations remain reproducible while giving you a rich historical record of your decision-making process.

## 1. Initializing a Project

Every BayesGrove analysis lives inside an explicitly initialized project handle. This handle manages state, cache, and decision logs inside a `.bayesgrove/` directory.

```{r setup}
library(bayesgrove)

# Use a temporary directory for this vignette
project_root <- file.path(tempdir(), "bg-quickstart")
unlink(project_root, recursive = TRUE)

handle <- bg_init(path = project_root, project_name = "Quickstart")
print(handle)
```

## 2. Defining the Workflow Structure

In BayesGrove, everything is a node. For this demo, let's register some mock executors to simulate the workflow.

```{r register}
# Register a simple data provider
bg_register_node_kind(handle, "data", executor = function(node, inputs) {
  # In a real workflow, this would read from disk or a database
  data.frame(x = rnorm(100), y = rnorm(100))
})

# Register a simple model fitter
bg_register_node_kind(handle, "model", executor = function(node, inputs) {
  # Extract the upstream data
  raw_data <- inputs[[1]]
  sprintf("Fitted model on %d rows of data!", nrow(raw_data))
})
```

Now we build the graph:

```{r build_graph}
n_data <- bg_add_node(handle, kind = "data", label = "Simulated Dataset")
n_model <- bg_add_node(handle, kind = "model", label = "Baseline Fit", inputs = n_data)
```

## 3. Decision Gates

Instead of just running the model directly, let's add a "Decision Gate" to force us to explicitly approve the data before moving on. This captures provenance.

```{r add_gate}
gate <- bg_add_gate(
  project = handle,
  from = n_data,
  to = n_model,
  prompt = "Does the simulated data look ready for modeling?",
  options = c("yes", "no")
)
```

If we try to plan the workflow now, the runner will see that the graph is blocked by an unresolved gate:

```{r plan_blocked}
plan <- bg_plan(handle)
print(plan$blocked)
print(bg_pending_gates(handle))
```

We must explicitly answer the gate and provide a rationale. This rationale is logged permanently.

```{r answer_gate}
bg_answer_gate(
  project = handle,
  id = gate$id,
  choice = "yes",
  rationale = "The simulated distributions match our expectations."
)
```

## 4. Execution

With the gate resolved, the path is unblocked. We can now run the workflow.

```{r execution}
run_result <- bg_run(handle, mode = "sync")
print(run_result$summary)
```

And that's it! You've successfully built a reproducible execution graph, gated it with an explicit decision, and executed it synchronously. 

## Next Steps

As you advance, you can use `bg_branch()` to create alternative model paths (e.g., trying a different prior without losing your baseline), or use `bg_invalidate()` to safely supersede outdated results.
