---
title: "Getting Started with BayesGrove"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with BayesGrove}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

**BayesGrove** is a graph-based Bayesian workflow orchestrator designed to track the iterative nature of model building. It separates your analysis into two distinct layers:
1. **The Execution Graph (DAG)**: Built using `dagriculture`, representing the step-by-step transformations and computations (data cleaning, compiling, fitting, diagnosing).
2. **The Decision Layer**: An annotation overlay that captures the provenance of your choices (why did you change the prior? why did you filter outliers?) via decision gates.

This separation ensures that your computations remain reproducible while giving you a rich historical record of your decision-making process. The current package also includes a lightweight workflow-protocol layer: executors can emit summaries, `bg_next_actions()` can derive obligations and suggested actions from those summaries, and callers can pass the resulting planner holds into `bg_plan()` without conflating them with structural graph blockers.

## 1. Initializing a Project

Every BayesGrove analysis lives inside an explicitly initialized project handle. This handle manages state, cache, and decision logs inside a `.bayesgrove/` directory.

```{r setup}
library(bayesgrove)

# Use a temporary directory for this vignette
project_root <- file.path(tempdir(), "bg-quickstart")
unlink(project_root, recursive = TRUE)

handle <- bg_init(
  path = project_root,
  project_name = "Quickstart",
  workflow_packs = list("bayesguide.default_bayesian")
)
print(handle)
```

## 2. Defining the Workflow Structure

In BayesGrove, everything is a node. For this demo, let's register some mock executors to simulate the workflow.

```{r register}
# Register a simple data provider
bg_register_node_kind(handle, "data", executor = function(node, inputs) {
  # In a real workflow, this would read from disk or a database
  data.frame(x = rnorm(100), y = rnorm(100))
})

# Register a simple model fitter
bg_register_node_kind(handle, "model", executor = function(node, inputs) {
  # Extract the upstream data
  raw_data <- inputs[[1]]
  list(
    result = sprintf("Fitted model on %d rows of data!", nrow(raw_data)),
    summaries = list(list(
      summary_kind = "optimizer_diagnostics",
      passed = FALSE,
      severity = "warning",
      metrics = list(max_gradient = 0.1)
    ))
  )
})

bg_register_node_kind(handle, "compare", executor = function(node, inputs) {
  sprintf("Compared downstream artifact: %s", inputs[[1]])
})
```

Now we build the graph:

```{r build_graph}
n_data <- bg_add_node(handle, kind = "data", label = "Simulated Dataset")
n_model <- bg_add_node(handle, kind = "model", label = "Baseline Fit", inputs = n_data)
n_compare <- bg_add_node(
  handle,
  kind = "compare",
  label = "Comparison Step",
  inputs = n_model
)
```

## 3. Decision Gates

Instead of just running the model directly, let's add a "Decision Gate" to force us to explicitly approve the data before moving on. This captures provenance.

```{r add_gate}
gate <- bg_add_gate(
  project = handle,
  from = n_data,
  to = n_model,
  prompt = "Does the simulated data look ready for modeling?",
  options = c("yes", "no")
)
```

If we try to plan the workflow now, the runner will see that the graph is blocked by an unresolved gate:

```{r plan_blocked}
plan <- bg_plan(handle)
print(plan$blocked)
print(bg_pending_gates(handle))
```

We must explicitly answer the gate and provide a rationale. This rationale is logged permanently.

```{r answer_gate}
bg_answer_gate(
  project = handle,
  id = gate$id,
  choice = "yes",
  rationale = "The simulated distributions match our expectations."
)
```

## 4. Execution

With the gate resolved, the path is unblocked. We now run only through the fit
node so that we can inspect the workflow protocol before scheduling downstream
comparison work.

```{r execution}
run_result <- bg_run(handle, targets = n_model, mode = "sync")
print(run_result$summary)
```

## 5. Workflow Guidance And External Holds

The fit executor emitted a lightweight warning summary. `bg_next_actions()`
turns that evidence into an obligation plus a matching next action:

```{r protocol}
next_steps <- bg_next_actions(handle)
next_steps$obligations[[1]]$kind
next_steps$actions[[1]]$payload$decision_type
```

Those blocking obligations can be passed back into planning as explicit
external holds. This keeps workflow guidance distinct from structural graph
state:

```{r external_holds}
held_plan <- bg_plan(
  handle,
  external_holds = next_steps$metadata$external_holds
)

held_plan$blocked
held_plan$external_blocked
```

In this example, the downstream comparison node is not structurally broken. It
is simply being held until the warning summary is reviewed or replaced by a
clean rerun.

## Next Steps

As you advance, you can use `bg_branch()` to create alternative model paths
(e.g., trying a different prior without losing your baseline), `bg_invalidate()`
to supersede outdated results, and `bg_next_actions()` plus `bg_plan(...,
external_holds = ...)` to keep workflow obligations visible during scheduling.
