---
title: "Primed Prior Case Studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Primed Prior Case Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Why this vignette exists

This vignette shows how to map the three case studies from Fazio, Scholz,
Aguilar, and Bürkner's _Primed Priors for Simulation-Based Validation of
Bayesian Models_ (<https://arxiv.org/abs/2408.06504>) into `bayesgrove`.

It is a workflow showcase, not a full numerical reproduction of the paper. The
heavy model fitting, SBC loops, and plotting are still delegated to user node
executors, while `bayesgrove` provides:

- explicit graph structure for priming, fitting, and SBC,
- branching over priors, priming datasets, and prior weights,
- decision logging for calibration review,
- optional workflow packs for prior review, model checks, model comparison,
  process guidance, model taxonomy, Stan diagnostics, and DAG-based causal
  checks.

The key idea is to treat a primed-prior workflow as a reusable graph pattern:

1. define a target model,
2. define one or more synthetic-data generators for priming,
3. fit a priming model or build a primed prior artifact,
4. run SBC under the primed prior,
5. review calibration explicitly,
6. compare branches across prior weights or priming regimes.

# Setup

```{r setup}
library(bayesgrove)

project_root <- file.path(tempdir(), "bg-primed-priors")
unlink(project_root, recursive = TRUE, force = TRUE)

handle <- bg_init(
  path = project_root,
  project_name = "primed-priors",
  workflow_packs = list(
    "bayesgrove.default_bayesian",
    "bayesgrove.process_guidance",
    "bayesgrove.prior_workflow",
    "bayesgrove.model_checks",
    "bayesgrove.model_selection",
    "bayesgrove.model_taxonomy",
    "bayesgrove.stan_workflow",
    "bayesgrove.causal_dagitty"
  )
)

vapply(bg_workflow_packs(handle), function(pack) pack$pack_id, character(1))
```

# A reusable primed-SBC graph shape

Across all three case studies, the graph structure is similar:

- one node (or branch) defines the model and prior configuration,
- one node builds the synthetic data used for priming,
- one node materializes the primed prior or preconditioning fit,
- one node generates SBC datasets,
- one node runs SBC diagnostics and emits `sbc_result` summaries,
- optional comparison nodes rank competing prior weights or priming schemes.

In practice, the node executors stay project-specific. The package-level
contract is only that they return plain R data and summary records.

```{r primed-sbc-template, eval = FALSE}
bg_register_node_kind(handle, "model_spec", executor = function(node, inputs) {
  node$params
})

bg_register_node_kind(handle, "prime_data", executor = function(node, inputs) {
  spec <- inputs[[1]]

  list(
    priming_data = node$params$generator(spec),
    strategy = node$params$strategy,
    prior_weight = node$params$prior_weight
  )
})

bg_register_node_kind(handle, "primed_prior", executor = function(node, inputs) {
  spec <- inputs[[1]]
  prime <- inputs[[2]]

  list(
    prior_artifact = node$params$build_prior(spec, prime),
    summaries = list(
      list(
        summary_kind = "prior_spec",
        passed = TRUE,
        severity = "ok",
        metrics = list(
          strategy = prime$strategy,
          prior_weight = prime$prior_weight
        )
      ),
      list(
        summary_kind = "prior_predictive_check",
        passed = TRUE,
        severity = "ok",
        metrics = node$params$prior_predictive_metrics(spec, prime)
      )
    )
  )
})

bg_register_node_kind(handle, "sbc_run", executor = function(node, inputs) {
  spec <- inputs[[1]]
  primed <- inputs[[2]]

  sbc <- node$params$run_sbc(spec, primed)

  list(
    sbc = sbc,
    summaries = list(list(
      summary_kind = "sbc_result",
      passed = sbc$passed,
      severity = if (sbc$passed) "ok" else "warning",
      metrics = sbc$metrics
    ))
  )
})
```

That template is enough to let the built-in packs ask for prior rationale,
prior predictive review, SBC review, process notes, and branch comparison.

# Case study 1: lower-bounded gamma regression

The paper's first case study uses a gamma regression on the `bodyfat` data with
a lower-bounded response and weak priors that can generate floating-point
underflow near zero. The practical lesson is that `bayesgrove` should treat the
priming strategy itself as a first-class branchable object.

## What to model in the graph

- the baseline gamma model,
- the priming generator based on fixed parameter values,
- the alternative priming generator based on external-data resampling,
- the prior weight `tau = M`,
- the SBC result and calibration review.

## A bayesgrove mapping

```{r gamma-case-study, eval = FALSE}
bg_register_node_kind(handle, "bodyfat_reference", executor = function(node, inputs) {
  node$params$data
})

bg_register_node_kind(handle, "gamma_prime_data", executor = function(node, inputs) {
  reference <- inputs[[1]]

  if (identical(node$params$strategy, "external_resample")) {
    idx <- sample.int(nrow(reference), size = node$params$M, replace = FALSE)
    priming_data <- reference[idx, , drop = FALSE]
  } else {
    priming_data <- node$params$simulate_gamma(
      M = node$params$M,
      beta_0 = 1,
      beta = rep(0.1, 13),
      alpha = 1
    )
  }

  list(
    priming_data = priming_data,
    strategy = node$params$strategy,
    prior_weight = node$params$M
  )
})

bg_register_node_kind(handle, "gamma_primed_prior", executor = function(node, inputs) {
  prime <- inputs[[2]]

  list(
    prior_weight = prime$prior_weight,
    summaries = list(
      list(
        summary_kind = "prior_spec",
        passed = TRUE,
        severity = "ok",
        metrics = list(strategy = prime$strategy, prior_weight = prime$prior_weight)
      ),
      list(
        summary_kind = "prior_predictive_check",
        passed = TRUE,
        severity = "ok",
        metrics = list(censoring_threshold = 1e-16, primed = TRUE)
      )
    )
  )
})

bg_register_node_kind(handle, "gamma_sbc", executor = function(node, inputs) {
  list(
    summaries = list(list(
      summary_kind = "sbc_result",
      passed = TRUE,
      severity = "ok",
      metrics = list(
        log_gamma = -5.2,
        M = node$params$M,
        prior_width = node$params$prior_width
      )
    ))
  )
})
```

## Why this graph is useful

The important workflow object is not only the fit, but also the decision about
how to generate priming data. In `bayesgrove`, that means you can:

- branch over prior widths,
- branch over `M`,
- branch over fixed-parameter versus resampled priming data,
- compare which branch yields acceptable `sbc_result` summaries.

The paper's heuristic `M ~= dim(theta)` maps naturally to a branch parameter and
to the `process_guidance` pack's explicit workflow-preflight note.

# Case study 2: R2D2 shrinkage prior

The second case study uses primed priors to make SBC feasible for an R2D2-style
shrinkage prior in high-dimensional regression. The paper studies two priming
regimes:

1. an intercept-only generator, and
2. a fixed-coefficients generator with a few large signals.

This is a good fit for `bayesgrove` because those two regimes are genuinely
different workflow branches.

## A bayesgrove mapping

```{r shrinkage-case-study, eval = FALSE}
bg_register_node_kind(handle, "shrinkage_scenario", executor = function(node, inputs) {
  node$params
})

bg_register_node_kind(handle, "r2d2_prime_data", executor = function(node, inputs) {
  scenario <- inputs[[1]]

  generator <- switch(
    node$params$scenario,
    intercept_only = function() rep(0, scenario$p),
    fixed_coefficients = function() {
      beta <- rep(0, scenario$p)
      beta[c(1:5, (scenario$p - 4):scenario$p)] <- sample(c(1e2, 1e3, 1e4), 10, replace = TRUE)
      beta
    }
  )

  list(
    beta = generator(),
    prior_weight = node$params$tau,
    p = scenario$p,
    n = scenario$n
  )
})

bg_register_node_kind(handle, "r2d2_sbc", executor = function(node, inputs) {
  prime <- inputs[[2]]

  list(
    summaries = list(list(
      summary_kind = "sbc_result",
      passed = TRUE,
      severity = "ok",
      metrics = list(
        scenario = node$params$scenario,
        tau = prime$prior_weight,
        p = prime$p,
        n = prime$n
      )
    ))
  )
})
```

## Why this graph is useful

The paper recommends treating `tau` as a tunable hyperparameter and tailoring
synthetic data to the intended sparsity regime. In `bayesgrove`, that means:

- create one branch for `intercept_only`,
- create another branch for `fixed_coefficients`,
- branch further across candidate `tau` values,
- use `bayesgrove.model_selection` to compare which branch produces the most
  stable SBC summaries.

This is also a good place to use `bayesgrove.model_taxonomy`: the workflow can
explicitly record that the branch is a `PAD` model with utility priorities
centered on `parameter_recoverability`, `robustness`, and `convergence`.

# Case study 3: latent mediation model

The third case study is the richest one for `bayesgrove`. It combines:

- latent-variable mediation structure,
- prior-predictive failure under weak defaults,
- partial priming of selected parameters,
- SBC review for a `latent_inference` goal,
- explicit causal structure.

## A bayesgrove mapping

```{r latent-case-study, eval = FALSE}
bg_register_node_kind(handle, "latent_mediation_spec", executor = function(node, inputs) {
  node$params
})

bg_register_node_kind(handle, "latent_prime_data", executor = function(node, inputs) {
  spec <- inputs[[1]]

  list(
    synthetic_data = node$params$simulate_latent_data(spec),
    partial_priming = TRUE,
    primed_parameters = c(
      "latent_means",
      "latent_scales",
      "structural_slopes"
    )
  )
})

bg_register_node_kind(handle, "latent_sbc", executor = function(node, inputs) {
  list(
    summaries = list(
      list(
        summary_kind = "prior_predictive_check",
        passed = TRUE,
        severity = "ok",
        metrics = list(target = "latent_mediation")
      ),
      list(
        summary_kind = "sbc_result",
        passed = TRUE,
        severity = "ok",
        metrics = list(M = 10, partial_priming = TRUE)
      )
    )
  )
})
```

The paper's practical move—treating some generated latent variables as observed
inside the priming step—maps naturally to a dedicated `latent_prime_data` node.
That node becomes a visible artifact in the workflow instead of a hidden part of
an SBC script.

## Adding the DAG layer with `bayesgrove.causal_dagitty`

For this mediation example, the new `bayesgrove.causal_dagitty` pack is a good
fit. It expects nodes that emit `dagitty_adjustment` and `dagitty_implications`
summaries. One minimal node set looks like this:

```{r dagitty-case-study, eval = FALSE}
bg_register_node_kind(handle, "dagitty_adjustment", executor = function(node, inputs) {
  if (!requireNamespace("dagitty", quietly = TRUE)) {
    cli::cli_abort("The {.pkg dagitty} package is required.")
  }

  dag <- dagitty::dagitty(node$params$dag_code)
  sets <- dagitty::adjustmentSets(
    dag,
    exposure = node$params$exposure,
    outcome = node$params$outcome
  )

  list(
    adjustment_sets = sets,
    summaries = list(list(
      summary_kind = "dagitty_adjustment",
      passed = TRUE,
      severity = "ok",
      metrics = list(
        n_sets = length(sets),
        exposure = node$params$exposure,
        outcome = node$params$outcome
      )
    ))
  )
})

bg_register_node_kind(handle, "dagitty_implications", executor = function(node, inputs) {
  if (!requireNamespace("dagitty", quietly = TRUE)) {
    cli::cli_abort("The {.pkg dagitty} package is required.")
  }

  dag <- dagitty::dagitty(node$params$dag_code)
  implications <- dagitty::impliedConditionalIndependencies(dag)

  list(
    implications = implications,
    summaries = list(list(
      summary_kind = "dagitty_implications",
      passed = TRUE,
      severity = "ok",
      metrics = list(n_implications = length(implications))
    ))
  )
})
```

With that node set in place, `bayesgrove.causal_dagitty` will:

- ask for a DAG-based adjustment review,
- ask for a review of the DAG's implied independencies,
- log both decisions in the branch history, and
- optionally consume a `causal_selection_contract` summary that locks required
  identification terms, excludes forbidden controls, and ranks admissible
  precision candidates for later formula revisions or projection-based
  reduction.

If you want a plotting layer, `ggdag::tidy_dagitty()` can sit on top of the same
`dag_code`, but `dagitty` should stay the semantic core.

For a primed-prior mediation workflow, a useful next node after DAG adjustment
is a `causal_selection_contract` node whose summary metrics look like:

```{r causal-selection-contract, eval = FALSE}
list(
  summary_kind = "causal_selection_contract",
  passed = TRUE,
  severity = "ok",
  metrics = list(
    required_terms = c("treatment", "pre_treatment_confounder"),
    forbidden_terms = c("post_treatment_proxy"),
    ranked_candidate_terms = c("baseline_risk", "age", "sex")
  )
)
```

Once reviewed, that contract can keep the branch's formulas or projpred-style
submodel search inside the DAG-approved candidate set.

# Recommended branch structure

For these case studies, a good pattern is:

1. keep one branch per priming regime,
2. keep one sub-branch per prior weight or prior family,
3. treat SBC outputs as summaries instead of ad hoc plots,
4. use `bg_record_decision()` to log whether calibration is acceptable,
5. only compare branches after the relevant prior, SBC, and DAG reviews are in
   place.

That gives you a workflow object that is much easier to audit than a single SBC
script.

# Closing remarks

The paper's main contribution is a way to move validation toward practically
relevant parts of parameter space. In `bayesgrove`, that maps cleanly to graph
structure:

- priming strategies become nodes,
- prior weights become branch parameters,
- SBC outputs become summaries,
- calibration judgments become explicit decisions,
- causal assumptions can become DAG-backed review steps.

That is the core pattern to reuse if you want to turn primed-prior validation
into a durable, inspectable workflow instead of a one-off analysis script.
