---
title: "A recorded model search"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A recorded model search}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The experimental research interface records candidates, evidence, and why an
investigation moved between them. R users, Glade, and agents can use the same
operations. The execution graph remains available for running computations.

This example uses explicitly illustrative evidence so it runs without Stan.
For a real fit and predictive check, run `Rscript tools/research-search-demo.R`
from the repository root with cmdstanr and CmdStan installed.

```text
Normal candidate
  ├─ Record a predictive check without creating a revision
  ├─ Revise P: lognormal candidate
       ├─ Record its predictive check
       └─ Close, compare, then reopen this path

Compare the two candidates, retaining the closed candidate's evidence.
```

## Start with a question

```{r}
library(bayesgrove)
h <- bg_init(tempfile("research-search-"))
human <- list(id = "researcher", type = "human")
agent <- list(id = "assistant", type = "agent")

review_rule <- list(
  id = "review-predictions",
  actions = "accept",
  type = "require_review",
  check = "ppc",
  message = "Review the predictive evidence before accepting this candidate.",
  interpretation = "Inspect the features that matter for the analysis question.",
  references = "https://arxiv.org/abs/2209.02439"
)

state <- bg_research_init(
  h,
  question = "How should we model these positive measurements?",
  goal = list(kind = "observable_prediction", focus = "positive support"),
  actor = human,
  mode = "guide",
  rules = list(review_rule)
)
```

A candidate specifies P (joint distribution), with optional A (approximator)
and D (training data). Specifications are plain data. The labels below are
illustrative; a reproducible analysis should also retain exact model source,
data, and settings, not only labels or mutable file paths.

```{r}
proposal <- bg_research_propose(
  h,
  action = list(
    kind = "create", label = "Normal",
    components = list(P = list(likelihood = "normal", prior = "illustrative"))
  ),
  actor = human,
  rationale = "Start with a simple candidate and inspect its predictions."
)
state <- bg_research_apply(h, proposal)
normal <- tail(state$history, 1)[[1]]$id
```

## Record evidence and revise

This local helper only shortens the example. Production clients can inspect the
proposal's `findings` and `allowed` before calling apply. Apply checks them again.

```{r}
record <- function(action, rationale, actor = human) {
  proposal <- bg_research_propose(h, action, actor, rationale)
  bg_research_apply(h, proposal)
}

state <- record(
  list(kind = "evidence", candidate_id = normal, label = "Illustrative PPC",
       check = "ppc", utility = "predictive_performance",
       result = list(source = "illustrative", concern = "negative predictions")),
  "The example check raises a concern about predictive support."
)
normal_evidence <- tail(state$history, 1)[[1]]$id

state <- record(
  list(kind = "revise", candidate_id = normal, label = "Lognormal",
       components = list(P = list(likelihood = "lognormal", prior = "illustrative")),
       basis = list(evidence_ids = normal_evidence)),
  "Investigate a likelihood with positive support."
)
lognormal <- tail(state$history, 1)[[1]]$id
```

A revision replaces the supplied components whole and inherits the others.
The parent and its evidence stay unchanged. Changing A or D uses the same
operation. A diagnostic or an interpretation adds evidence rather than a revision.

```{r}
state <- record(
  list(kind = "evidence", candidate_id = lognormal, label = "Illustrative PPC",
       check = "ppc", utility = "predictive_performance",
       result = list(source = "illustrative", concern = "none in this example")),
  "Record the alternative's check without treating it as proof of adequacy.",
  actor = agent
)
lognormal_evidence <- tail(state$history, 1)[[1]]$id

state <- record(
  list(kind = "close", candidate_id = lognormal),
  "Pause this path while considering other alternatives."
)
state <- record(
  list(kind = "compare", candidate_ids = c(normal, lognormal),
       evidence_ids = c(normal_evidence, lognormal_evidence),
       criteria = "Predictive support in the illustrative checks",
       result = list(preferred_for_support = lognormal)),
  "Compare both candidates; the closed path's evidence remains available."
)
comparison <- tail(state$history, 1)[[1]]$id

state <- record(list(kind = "reopen", candidate_id = lognormal,
                    basis = list(comparison_ids = comparison)),
                "Resume this path after the comparison.")
```

Closing a lineage prevents revisions, evidence collection, and acceptance below
it until reopening. Closing does not delete anything or automatically reject a
candidate. Comparisons can include candidates from any lineage.

## Switch intervention without rewriting history

Record mode preserves actions. Guide adds findings and explanations. Enforce
also holds actions with unmet rules. Structural checks apply in every mode.

```{r}
state <- record(list(kind = "configure", mode = "enforce", rules = list(review_rule)),
                "Require human review before recording acceptance.")
acceptance <- list(kind = "accept", candidate_id = lognormal,
                   evidence_ids = lognormal_evidence,
                   basis = list(comparison_ids = comparison))
proposal <- bg_research_propose(h, acceptance, agent, "Select the alternative.")
proposal$allowed
proposal$findings[[1]]$message
```

An agent can record a review, but it cannot satisfy this rule's human-review
requirement. A human review covers only the evidence it cites. New evidence or
a new candidate needs its own review.

```{r}
state <- record(list(kind = "review", candidate_id = lognormal,
                    evidence_ids = lognormal_evidence),
                "For this example, acknowledge the support check and its limited scope.")
proposal <- bg_research_propose(h, acceptance, human,
                              "Accept for the example's stated goal, subject to further checks.")
state <- bg_research_apply(h, proposal)
state$version
```

A proposal becomes stale after any successful research mutation. Read the state
and propose again rather than replaying it. Declared actor identity is provenance,
not authentication; a host must establish identity for untrusted callers.
Enforce applies to this research interface, not arbitrary R code or legacy
`bg_run()` calls. Existing workflow packs retain their separate behavior.

## Reframe without losing earlier decisions

A declared human actor can correct the question or change the inferential goal.
Earlier records keep their `goal_version`; their reviews do not authorize
acceptance under the new goal. The evidence itself remains available.

```{r}
state <- record(
  list(kind = "reframe", question = "How well are the largest measurements predicted?",
       goal = list(kind = "observable_prediction", focus = "upper tail")),
  "Focus the investigation on extreme observations."
)
proposal <- bg_research_propose(h, acceptance, human,
                              "Reconsider the candidate for the revised goal.")
proposal$allowed
```

## Reopen and share

```{r}
path <- h@path
bg_close(h)
h <- bg_open(path)
stopifnot(identical(bg_research_state(h), state))
bg_close(h)
```

`bg_snapshot()`, `bg_bundle()`, and `bg_export_report()` include research state.
The [research-search contract](https://github.com/sims1253/bayesgrove/blob/master/architecture/research-search.md)
lists all action fields and declarative rule types, including specification
constraints and evidence requirements. There is no built-in universal model
score: comparisons state their criteria and retain their evidence.
