---
title: "Reproducible research with bayesgrove"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article covers archiving an analysis, verifying its records, and
describing it in a methods section. See `vignette("concepts")` for definitions
and `vignette("eight-schools")` for a worked analysis.

# What the bundle contains

`bg_bundle()` produces a single portable `.tar.gz` of the project:

```{r}
library(bayesgrove)

handle <- bg_open("analysis/")
bundle <- bg_bundle(handle, path = "analysis-bundle.tar.gz")
```

The archive contains these files under `.bayesgrove/`:

| File | Contents |
|:-----|:---------|
| `graph/graph.json` | The full analysis graph: nodes, edges, params, gates. |
| `decisions/decisions.jsonl` | Every recorded decision: prompt, choice, rationale, scope, evidence, timestamps. |
| `workflow/summaries.jsonl` | Typed diagnostic evidence (HMC diagnostics, PPC p-values, LOO Pareto-k counts, LOO-PIT calibration, SBC ranks) with severities and execution fingerprints. |
| `runs/jobs.jsonl` | The execution history: which node ran when, with what outcome. |
| `cache/` | Content-addressed artifacts (large fit objects are excluded unless `include_fits = TRUE`). |
| `bundle_manifest.json` | A reproducibility manifest: R version, platform, package versions, CmdStan version, creation time. |

Data attached with `bg_set_node_data()` are included, even before execution.
External files referenced by node params, such as Stan programs, are not copied
or relocated. Archive those files separately and update their paths before
rerunning on another machine.

Decisions and summaries are appended to JSONL logs with increasing sequence
numbers. Entries are not edited in place, so the log retains earlier decisions
when a prior or model changes.

Each summary contains its node's execution fingerprint: a deterministic hash
of the params, upstream fingerprints, Stan program contents, and environment.
Use it to check whether the evidence cited by a decision matches the reported
configuration.

# Restoring a bundle on another machine

Untar the bundle and open it:

```{r}
utils::untar("analysis-bundle.tar.gz", exdir = "restored")
handle <- bg_open("restored/analysis")
```

`bg_open()` compares the bundle's reproducibility manifest against the
current session and warns when the R version, platform, or CmdStan version
differ. This warning does not invalidate cached artifacts; fingerprints
determine what reruns. Matching the recorded environment may be necessary
for bit-level reproduction.

Opening a project does not evaluate saved user executor source. Built-in
executors restore by reference. Review saved source before calling
`bg_restore_executors(trust = TRUE)`, or re-register custom executors with
`bg_register_node_kind()`.

# Read the decision and summary logs

Both logs are plain JSONL and parse with any JSON reader:

```{r}
decisions <- lapply(
  readLines("restored/analysis/.bayesgrove/decisions/decisions.jsonl"),
  jsonlite::fromJSON
)
summaries <- lapply(
  readLines("restored/analysis/.bayesgrove/workflow/summaries.jsonl"),
  jsonlite::fromJSON
)
```

Use the logs to check, for example, whether every
`hmc_diagnostics` summary with severity `warning` or `error` is referenced
by at least one `computation_review` decision.

For a human-readable rendering, `bg_export_report()` produces a Markdown or
HTML report with the graph as a Mermaid diagram, the decision history, and
the artifact index.

# SBC and simulation studies at scale

The wave scheduler dispatches independent nodes concurrently when their
inputs are ready. Start `mirai` daemons before running the graph:

```{r}
mirai::daemons(8)
bg_run(handle) # sbc / simulation nodes fan out across the daemons
mirai::daemons(0)
```

The built-in `sbc` node kind runs the Talts et al. (2018) rank-statistic
loop and emits an `sbc_result` summary containing a rank histogram and a
chi-squared uniformity check. The SBC review pack uses this summary to request
reviews. Input comes from an upstream generator executor as plain data:
`list(simulations = list(list(theta = ..., data = ...), ...))`.
Generator code requires the same explicit trust as other custom executors.
The executor thins each fit to a common number bounded by
bulk ESS before ranking, and refuses to grade histograms whose expected bin
counts are below five. Twenty simulations are a smoke test; use hundreds for a
substantive calibration study.

Parallel workers always preserve the content-addressed R fit artifact. Raw
sampler CSV paths can point into daemon-local temporary directories, however,
so `bg_run()` warns when a parallel wave contains fit nodes. Use
`parallel = "never"` when durable raw CSV files are required. See `?bg_run` for
the wave semantics and `vignette("simulation-study")` for a full design.

# Reporting template

A "data and code availability" statement for a bayesgrove-backed analysis:

> We used bayesgrove (Scholz, YEAR) to record the model graph, diagnostic
> evidence, and analysis decisions in machine-readable form. The project
> bundle is archived at DOI. It contains the decision log with rationales,
> typed diagnostic summaries with execution fingerprints, and a
> reproducibility manifest (R X.Y.Z, CmdStan X.Y.Z).

Cite the package and its workflow reference with
`citation("bayesgrove")`.
