---
title: "Reproducible research with bayesgrove"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article is for researchers who want to *report* a bayesgrove-backed
analysis: what to archive, what a reviewer can verify from the archive, and
how to phrase it in a methods section. The companion narrative is
`vignette("concepts")`; the mechanics of a full analysis are in
`vignette("eight-schools")`.

# What the bundle contains

`bg_bundle()` produces a single portable `.tar.gz` of the project:

```{r}
library(bayesgrove)

handle <- bg_open("analysis/")
bundle <- bg_bundle(handle, path = "analysis-bundle.tar.gz")
```

Inside the archive, under `.bayesgrove/`, an independent reader finds:

| File | Contents |
|:-----|:---------|
| `graph/graph.json` | The full analysis graph: nodes, edges, params, gates. |
| `decisions/decisions.jsonl` | Every recorded decision: prompt, choice, rationale, scope, evidence, timestamps. |
| `workflow/summaries.jsonl` | Typed diagnostic evidence (HMC diagnostics, PPC p-values, LOO Pareto-k counts, LOO-PIT calibration, SBC ranks) with severities and execution fingerprints. |
| `runs/jobs.jsonl` | The execution history: which node ran when, with what outcome. |
| `cache/` | Content-addressed artifacts (large fit objects are excluded unless `include_fits = TRUE`). |
| `bundle_manifest.json` | A reproducibility manifest: R version, platform, package versions, CmdStan version, creation time. |

Two properties make this an audit trail rather than a snapshot:

1. **Append-only logs.** Decisions and summaries are JSONL appends with
   monotonic sequence numbers; nothing is edited in place. The history of a
   revised prior is the sequence of decisions, not the final state.
2. **Fingerprints.** Every summary carries the execution fingerprint of the
   node that produced it — a deterministic hash of the node's params, its
   upstream fingerprints, the Stan program *contents*, and the environment.
   A reader can verify that the diagnostic evidence cited for a decision
   came from the exact configuration that was eventually reported.

# Restoring a bundle on another machine

Untar the bundle and open it:

```{r}
utils::untar("analysis-bundle.tar.gz", exdir = "restored")
handle <- bg_open("restored/analysis")
```

`bg_open()` compares the bundle's reproducibility manifest against the
current session and warns when the R version, platform, or CmdStan version
differ. Cached artifacts stay usable either way — fingerprints, not the
manifest, decide what reruns — but the warning tells you up front that
bit-level reproduction may require matching the recorded environment.

Executors are **never** restored implicitly: opening a project runs no
project-supplied code. Built-in executors re-register with
`bg_use_cmdstanr()` / `bg_use_brms()`; user executors restore only through
the explicit `bg_restore_executors(trust = TRUE)` gate after you have
reviewed their persisted source.

# The machine-readable protocol trace

Both logs are plain JSONL and parse with any JSON reader:

```{r}
decisions <- lapply(
  readLines("restored/analysis/.bayesgrove/decisions/decisions.jsonl"),
  jsonlite::fromJSON
)
summaries <- lapply(
  readLines("restored/analysis/.bayesgrove/workflow/summaries.jsonl"),
  jsonlite::fromJSON
)
```

This supports checks that a notebook-based workflow cannot: e.g. "every
`hmc_diagnostics` summary with severity `warning` or `error` is referenced
by at least one `computation_review` decision". That is the property
bayesgrove *enforces* during the analysis (blocking obligations hold
downstream nodes), and the trace lets a reviewer confirm it after the fact.

For a human-readable rendering, `bg_export_report()` produces a Markdown or
HTML report with the graph (as a Mermaid diagram), the full decision
provenance, and the artifact index.

# SBC and simulation studies at scale

Calibration studies are hundreds of independent fits — exactly the shape the
wave scheduler parallelizes. Start `mirai` daemons and run; nodes whose
inputs are ready dispatch concurrently, and results are byte-identical to a
sequential run because artifacts are content-addressed:

```{r}
mirai::daemons(8)
bg_run(handle) # sbc / simulation nodes fan out across the daemons
mirai::daemons(0)
```

The built-in `sbc` node kind runs the Talts et al. (2018) rank-statistic
loop and emits an `sbc_result` summary (rank histogram plus a chi-squared
uniformity check) that the SBC review pack turns into an obligation like any
other piece of evidence. Its input is plain data from an upstream generator
executor—`list(simulations = list(list(theta = ..., data = ...), ...))`—so
generator code follows the same explicit executor trust gate as other
project-defined code. The executor thins each fit to a common number bounded by
bulk ESS before ranking, and refuses to grade histograms whose expected bin
counts are below five. Twenty simulations are a smoke test; use hundreds for a
substantive calibration study.

Parallel workers always preserve the content-addressed R fit artifact. Raw
sampler CSV paths can point into daemon-local temporary directories, however,
so `bg_run()` warns when a parallel wave contains fit nodes. Use
`parallel = "never"` when durable raw CSV files are required. See `?bg_run` for
the wave semantics and `vignette("simulation-study")` for a full design.

# Reporting template

A "data and code availability" statement for a bayesgrove-backed analysis:

> The analysis was managed with bayesgrove (Scholz, YEAR), which records
> the model graph, all diagnostic evidence, and all analyst decisions in
> machine-readable form. The complete project bundle — including the
> decision log with rationales, typed diagnostic summaries with execution
> fingerprints, and a reproducibility manifest (R X.Y.Z, CmdStan X.Y.Z) —
> is archived at DOI. Model fits can be regenerated from the bundle with
> `bg_open()` followed by `bg_run()`; environment mismatches with the
> recorded manifest are surfaced on open.

Cite the package and the workflow paper it operationalizes with
`citation("bayesgrove")`.
