---
title: "Teaching with bayesgrove"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article is for instructors running a Bayesian modeling course on top of
bayesgrove's decision log. The pedagogical claim is simple: students learn
the workflow loop (Gelman et al. 2020) faster when the software refuses to
let them skip its steps. A fit with divergences cannot silently feed a
model comparison; the student must look at the evidence and record a
judgment, and that judgment is what you grade.

# The assignment shape

Students submit two files:

1. The exported report — `bg_export_report(handle, format = "html")` — the
   human-readable narrative with the graph diagram and decision provenance.
2. The project bundle — `bg_bundle(handle)` — the machine-readable trail you
   verify against the report.

Because the bundle's decision and summary logs are append-only and every
summary carries the fingerprint of the configuration that produced it, the
report cannot claim a clean fit the logs do not support. Verifying a
submission is opening the bundle, not re-running it:

```{r}
utils::untar("student-submission.tar.gz", exdir = "check")
handle <- bg_open("check/project")

bg_status(handle) # blocked? then obligations were left unresolved
bg_next_actions(handle) # prints outstanding obligations as a checklist
bg_export_report(handle, format = "md") # regenerate and diff against the submitted report
```

# Grading checklist: obligations to rubric items

The workflow packs raise obligations; each maps to something gradeable.

| Obligation (pack) | Evidence behind it | Rubric item |
|:------------------|:-------------------|:------------|
| `review_computation_validity` | `hmc_diagnostics` summary with warning/error severity | Every sampler warning has a recorded `computation_review` decision whose rationale names the diagnostic (divergences, R-hat, ESS, E-BFMI). |
| Prior rationale (prior pack) | Prior/prior-predictive nodes | Every prior choice has a decision with a substantive rationale, not "default". |
| Prior/posterior predictive review | `prior_predictive_check`, `posterior_predictive_check` summaries | PPC results interpreted, not just produced; extreme p-values acknowledged. |
| Calibration review (SBC pack) | `sbc_result`, `loo_pit_calibration` summaries | Calibration checked where the assignment requires it; miscalibration discussed. |
| `review_fit_criticism` | Branch-scoped diagnostic problems | Model revisions happen on branches with the criticism recorded, not by silently overwriting params. |
| `compare_candidate_branches` / `accept_or_reject_branch` | `comparison_results`, `stacking_weights` summaries | The final model was *chosen*: a comparison exists and a branch-disposition decision cites it. |

Two log-level checks catch the common shortcuts:

- **No orphaned warnings.** Every summary in `workflow/summaries.jsonl` with
  severity `warning`/`error` should be referenced (via its node or evidence
  field) by some decision in `decisions/decisions.jsonl`.
- **Rationales are text, not tokens.** Sort decisions by rationale length;
  one-word rationales cluster at the bottom and are easy to spot-check.

# Course setup tips

- Have students start from `bg_init()` + `bg_use_cmdstanr()` +
  `bg_use_default_workflow()` so the baseline review pack is active from the
  first fit. Stricter packs (prior rationale, SBC review) can be introduced
  per assignment via `bg_use_workflow_packs()`.
- `bg_add_gate()` lets *you* author checkpoints into a template project —
  e.g. a gate before the hierarchical model that asks "justify the pooling
  structure" — distinct from the evidence-driven obligations
  (see `?bayesgrove-concepts` for the gates-vs-obligations table).
- The REPL (`bg_repl()`) shows the obligation checklist interactively and
  `explain <n>` prints an obligation's *why* together with its literature
  references — the teaching hook for "why does the software care about
  divergences?".
- Plots are one call per node: `bg_plot(handle, node_id)` renders PPC
  density overlays, LOO-PIT ECDFs, SBC rank histograms, and trace plots for
  flagged fits.

# What this does not automate

The log proves that decisions were recorded, timestamped, and tied to
evidence; it cannot prove they are *good*. Grading the quality of a
rationale — whether "heavier tails to absorb outliers" is justified for the
data at hand — remains the instructor's job. bayesgrove's contribution is
that the rationale exists, is attached to the exact evidence that prompted
it, and cannot be fabricated after the fact.
