---
title: "Teaching with bayesgrove"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Use bayesgrove's decision log to review students' modeling choices and their
reasons. Blocking obligations require students to record a review before
downstream computation can proceed.

# Assignment

Students submit two files:

1. A report from `bg_export_report(handle, format = "html")`, containing the
   graph diagram and decision history.
2. A project bundle from `bg_bundle(handle)`, containing the records used to
   verify the report.

Open the bundle to compare its decision and summary logs with the report:

```{r}
utils::untar("student-submission.tar.gz", exdir = "check")
handle <- bg_open("check/project")

bg_status(handle) # blocked? then obligations were left unresolved
bg_next_actions(handle) # prints outstanding obligations as a checklist
bg_export_report(handle, format = "md") # regenerate and diff against the submitted report
```

# Grading checklist

Use the obligations and their evidence to assess the analysis.

| Obligation (pack) | Evidence behind it | Rubric item |
|:------------------|:-------------------|:------------|
| `review_computation_validity` | `hmc_diagnostics` summary with warning/error severity | Every sampler warning has a recorded `computation_review` decision whose rationale names the diagnostic (divergences, R-hat, ESS, E-BFMI). |
| Prior rationale (prior pack) | Prior/prior-predictive nodes | Every prior choice has a decision with a substantive rationale, not "default". |
| Prior/posterior predictive review | `prior_predictive_check`, `posterior_predictive_check` summaries | PPC results interpreted, not just produced; extreme p-values acknowledged. |
| Calibration review (SBC pack) | `sbc_result`, `loo_pit_calibration` summaries | Calibration checked where the assignment requires it; miscalibration discussed. |
| `review_fit_criticism` | Branch-scoped diagnostic problems | Revisions use branches that retain the earlier model and record its criticism. |
| `compare_candidate_branches` / `accept_or_reject_branch` | `comparison_results`, `stacking_weights` summaries | A comparison supports the final model choice, and a branch-disposition decision cites it. |

Check the logs for missing reviews and incomplete rationales:

- Every summary in `workflow/summaries.jsonl` with
  severity `warning`/`error` should be referenced (via its node or evidence
  field) by some decision in `decisions/decisions.jsonl`.
- Sort decisions by rationale length to find one-word answers for closer review.

# Course setup tips

- Have students start from `bg_init()` + `bg_use_cmdstanr()` +
  `bg_use_default_workflow()` so the baseline review pack is active from the
  first fit. Stricter packs (prior rationale, SBC review) can be introduced
  per assignment via `bg_use_workflow_packs()`.
- Use `bg_add_gate()` to add checkpoints to a template project, such as a
  gate before a hierarchical model asking "justify the pooling structure".
  These differ from obligations raised by diagnostic evidence
  (see `?bayesgrove-concepts` for the gates-vs-obligations table).
- The REPL (`bg_repl()`) shows the obligation checklist interactively and
  `explain <n>` prints its reason and literature references. Use these to
  explain checks such as divergence diagnostics.
- Plots are one call per node: `bg_plot(handle, node_id)` renders PPC
  density overlays, LOO-PIT ECDFs, SBC rank histograms, and trace plots for
  flagged fits.

# Assess the reasoning

The log records decisions, timestamps, and evidence references. The instructor
must assess whether each rationale, such as "heavier tails to absorb outliers",
is justified for the data.
